---
title: "Creating a Metacheck Report"
vignette: >
  %\VignetteIndexEntry{Creating a Metacheck Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
library(metacheck)
```

So far we have looked at individual modules in isolation. In practice you usually want to run *several* checks on a paper at once and collect the results into a single document you can read, save, or hand to a co-author. This chapter shows how to go from a single paper object to a full Metacheck report.

We start with the all-in-one `report()` function, then unpack what it does — running one module with `module_run()`, running several with `report_module_run()`, and turning the result into a document with `report_qmd()`. Finally we cover the slightly tricky part: passing settings through to the modules.

## Loading a paper

Every report starts from a **paper object**, created with `read()` (see [Reading in a Paper](reading-a-paper.qmd)). Throughout this chapter we use the bundled demo paper, which `demopaper()` returns directly so the examples run without a network connection:

```{r}
paper <- demopaper()
paper$paper_id
```

In your own work you would instead read a converted file:

```{r}
#| eval: false
paper <- read("converted/my_manuscript.json")
```

Everything below works on this single `paper` object. The same functions also accept a *paperlist* (a list of papers), so the same code scales to a whole corpus — see [Batch processing](reading-a-paper.qmd) and [Using the Paper Database Corpora](papers-databases.qmd).

## The quick way: `report()`

`report()` runs a set of modules on a paper and writes a finished report to disk. With no other arguments it runs the **default battery** of sixteen modules and saves an HTML file named after the paper. The defaults span the whole manuscript: preregistration ([`prereg_check`](mod-prereg-check.qmd)), funding and conflict-of-interest statements ([`funding_check`](mod-funding-check.qmd), [`coi_check`](mod-coi-check.qmd)), statistical power ([`power`](mod-power.qmd)), data and code availability ([`repo_check`](mod-repo-check.qmd), [`code_check`](mod-code-check.qmd)), the reporting of statistics ([`stat_check`](mod-stat-check.qmd), [`stat_p_exact`](mod-stat-p-exact.qmd), [`stat_p_nonsig`](mod-stat-p-nonsig.qmd), [`stat_effect_size`](mod-stat-effect-size.qmd), [`marginal`](mod-marginal.qmd)), and the reference list ([`ref_accuracy`](mod-ref-accuracy.qmd), [`ref_replication`](mod-ref-replication.qmd), [`ref_retraction`](mod-ref-retraction.qmd), [`ref_pubpeer`](mod-ref-pubpeer.qmd), [`ref_summary`](mod-ref-summary.qmd)).

```{r}
#| eval: false
report(paper)
#> "[paper_id]_report.html"
```

The function returns the path to the file it wrote. By default that is `"<paper_id>_report.html"` in the working directory; set `output_file` to choose your own name and location.

### Choosing modules

You rarely need all the default checks. Pass a `modules` vector to run exactly the ones you want, in the order you want them reported:

```{r}
#| eval: false
report(paper,
       modules = c("stat_p_exact", "stat_p_nonsig", "marginal"))
```

Use `module_list()` to see every built-in module, or `module_help("module_name")` for the details of one. Custom modules are given by file path instead of name (see [Creating Modules](creating-modules.qmd)).

### Choosing the output format

`report()` can write either a rendered **HTML** file or the intermediate **Quarto** source (`qmd`). The `qmd` option is useful when you want to edit the report before rendering, or render it yourself with different options:

```{r}
#| eval: false
report(paper,
       modules = c("stat_p_exact", "marginal"),
       output_file = "demo_report.html",
       output_format = "html")   # the default

report(paper,
       output_file = "demo_report.qmd",
       output_format = "qmd")    # write the Quarto source instead
```

Rendering to HTML requires [Quarto](https://quarto.org) to be installed. If rendering fails for any reason, `report()` falls back to saving the `qmd` source (and warns you), so you never lose the result of running the modules.

## Running a single module

`report()` is convenient, but it always writes a file. While you are exploring a paper interactively it is often handier to run one module and look at the result in the console. That is what `module_run()` does:

```{r}
mo <- module_run(paper, "stat_p_exact")
```

The object it returns holds everything the module found. The pieces you will use most are the traffic light, the one-line summary, and the full results table:

```{r}
mo$traffic_light
mo$summary_text
head(mo$table) |>
  knitr::kable()
```

Printing the object gives a compact summary, and `module_report()` formats the same output as the Markdown section that would appear in a report:

```{r}
#| results: asis
module_report(mo) |> cat()
```

## Running multiple modules

To run several modules and keep their combined output *without* writing a file, use `report_module_run()`. It takes the same `modules` vector as `report()` and returns a list of module outputs:

```{r}
module_output <- report_module_run(
  paper,
  modules = c("stat_p_exact", "stat_p_nonsig", "marginal")
)
```

The result is a named list — one entry per module — that you can inspect individually:

```{r}
names(module_output)
module_output[["marginal"]]$traffic_light
module_output[["marginal"]]$summary_text
```

This is the same intermediate object `report()` builds internally. To turn it into a finished document yourself, pass it to `report_qmd()`, which produces the Quarto report text (header, summary section, and one section per module):

```{r}
#| eval: false
report_text <- report_qmd(module_output, paper)
writeLines(report_text, "demo_report.qmd")
```

Splitting the work this way — run the modules once with `report_module_run()`, then format with `report_qmd()` — is useful when the modules are slow (for example reference checks that query external services) and you want to render the report more than once without re-running them.

## Passing settings to modules

This is the part that trips people up. Some modules take **extra arguments** beyond `paper`. For example:

- `power(paper, seed = 8675309)` — the random seed used when sampling text for the LLM.
- `ref_replication(paper, show_outcomes = FALSE)` — whether to include replication outcomes.
- `repo_check(paper, local_path = NULL, local_only = FALSE)` — where to find the repository, and whether to skip online lookups.

`module_help("module_name")` shows the arguments a given module accepts:

```{r}
module_help("power")
```

How you supply these arguments depends on which function you are using.

### One module: pass arguments directly

With `module_run()`, extra arguments go straight through the `...`, exactly as if you were calling the module function itself:

```{r}
#| eval: false
module_run(paper, "power", seed = 123)
module_run(paper, "ref_replication", show_outcomes = TRUE)
```

### Several modules: the `args` list

When you run *several* modules at once with `report()` or `report_module_run()`, you cannot use `...`, because an argument like `seed` would be ambiguous — which module is it for? Instead you pass a single `args` argument: **a named list of lists**, where each name matches a module name and each value is the list of arguments for that module.

```{r}
#| eval: false
args <- list(
  power           = list(seed = 123),
  ref_replication = list(show_outcomes = TRUE),
  repo_check      = list(local_only = TRUE)
)

report_module_run(
  paper,
  modules = c("power", "ref_replication", "repo_check"),
  args = args
)
```

The same `args` list works with `report()`:

```{r}
#| eval: false
report(
  paper,
  modules = c("power", "ref_replication", "repo_check"),
  args    = args
)
```

A few rules make this easier to get right:

- **You only list the modules that need arguments.** Modules you do not mention in `args` run with their defaults, even if they are in `modules`. So a module taking no special settings (like `stat_p_exact`) never needs an entry.
- **The names must match the module names** in your `modules` vector. A mismatched name (a typo, or a module not actually being run) is silently ignored — its settings simply have no effect.
- **Each value is itself a list**, even when there is a single argument: `power = list(seed = 123)`, not `power = 123`.

Putting it together, here is a complete report that runs four modules, two of them with custom settings and two with their defaults:

```{r}
#| eval: false
report(
  paper,
  modules = c("stat_p_exact", "marginal", "power", "ref_replication"),
  args = list(
    power           = list(seed = 123),
    ref_replication = list(show_outcomes = TRUE)
  ),
  output_file = "demo_report.html"
)
```

## Summary

| You want to… | Use |
|---|---|
| Run modules and write a finished report file | `report(paper, modules, output_file, output_format, args)` |
| Run one module and inspect the result | `module_run(paper, module, ...)` |
| Run several modules, keep the result, write no file | `report_module_run(paper, modules, args)` |
| Format module output as report text | `report_qmd(module_output, paper)` or `module_report(mo)` |
| Pass settings to one module | extra arguments to `module_run()` |
| Pass settings to several modules | the `args` named-list-of-lists |
