---
title: "Your First Check"
vignette: >
  %\VignetteIndexEntry{Your First Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
library(metacheck)
```

This chapter does not explain anything about how Metacheck works. It just walks through the four things you will do most often, using two short example papers, so that you have something working before you read about how any of it works. If you have never used R before, that is fine: every line below is explained, and you can copy and paste each code chunk into the R console one at a time.

We will use two small manuscripts, both about the same simple study (whether people are slower to name the ink color of a word when the word itself names a different color — the "Stroop effect"):

- A **good** paper: <https://github.com/Lakens/stroop_effect_paper>, which reports its statistics correctly and shares its data and code openly.
- A **bad** paper: <https://github.com/Lakens/stroop_effect_paper_bad>, a version of the same study with a number of common mistakes deliberately introduced, so you can see what Metacheck flags when something is wrong.

## Step 1: Read in a paper

Metacheck works with PDF files. The function `convert()` turns a PDF into a structured format Metacheck can read, and `read()` loads that structured format into R as a **paper object** — everything Metacheck does, in every later chapter, starts from a paper object like this one.

Since our example papers live on GitHub, we first download the PDF with `download.file()`, a base R function that needs no extra packages:

```{r}
#| eval: false
download.file(
  "https://raw.githubusercontent.com/Lakens/stroop_effect_paper/main/output/analysis.pdf",
  destfile = "good_paper.pdf", mode = "wb"
)
```

```{r}
#| include: false
# a local copy is used here so this chapter does not depend on a live
# internet connection or an external conversion server when the book is
# built
good_paper_file <- "../converted/good_paper.json"
```

Now convert the PDF and read it in:

```{r}
#| eval: false
json_file <- convert(file_path = "good_paper.pdf", save_path = ".")
paper <- read(json_file)
```

```{r}
#| include: false
paper <- read(good_paper_file)
```

`convert()` sends the PDF to a free conversion server and gets back a structured file; this needs an internet connection and takes a few seconds. You only need to do this once per paper — after that, you can `read()` the saved file directly, without converting again.

::: callout-note
`convert()` defaults to a server at Eindhoven University of Technology that does not store your manuscript. See [Reading in a Paper](reading-a-paper.qmd) if you would rather run your own, fully local conversion server.
:::

## Step 2: Run one check

Metacheck organises its checks into **modules**. Each module looks for one specific thing. `module_run()` runs a single module on a paper object. Just printing the result shows a short report:

```{r}
#| results: asis
module_run(paper, "stat_effect_size")
```

`stat_effect_size` checks whether a reported statistical test (like a *t*-test) is accompanied by an effect size, and whether that effect size is actually consistent with the test. Every module gives a **traffic light** — green (no issues found), yellow (worth a closer look), red (a likely problem), or "na" (nothing for this module to check) — plus a short summary. The good paper reports *t*(130) = 18.04 together with Cohen's *d~z~* = 1.58, and that effect size is indeed what you would expect from a test statistic that size — hence the green light above.

Now download the bad paper's PDF the same way as in Step 1, and run the same check on it:

```{r}
#| eval: false
download.file(
  "https://raw.githubusercontent.com/Lakens/stroop_effect_paper_bad/main/analysis.pdf",
  destfile = "bad_paper.pdf", mode = "wb"
)
json_file <- convert(file_path = "bad_paper.pdf", save_path = ".")
bad_paper <- read(json_file)
```

```{r}
#| include: false
bad_paper <- read("../converted/bad_paper.json")
```

```{r}
#| results: asis
module_run(bad_paper, "stat_effect_size")
```

This comes back red: it reports the same *t*-test, *t*(130) = 18.04, but with a Cohen's *d~z~* of 0.25 — a value nowhere near what that test statistic implies. Metacheck does not know which number was typed wrong, but it correctly flags that the two do not agree, which is exactly the kind of inconsistency a reader re-typing numbers by hand is likely to miss.

### Getting specific pieces of information out

Printing the result is useful for reading, but if you want to use a piece of it elsewhere — say, to check many papers automatically — assign the result to a variable first, and then pull out just the part you need with `$`:

```{r}
mo_bad <- module_run(bad_paper, "stat_effect_size")
mo_bad$traffic_light
mo_bad$summary_text
```

`mo_bad$traffic_light` gives you just the colour, and `mo_bad$summary_text` gives you just the plain-language summary — the same two pieces of text that were part of the printed report above. Every module's result has more than this; see [Running the module](mod-power.qmd) in the Modules part of this book for what else is in there.

You now know the single most important function in Metacheck: `module_run(paper, "module_name")`. Every module in the [Modules](mod-power.qmd) part of this book works exactly this way — only the module name changes.

## Step 3: Create a full report

Most of the time you will not want to run modules one at a time — you want every relevant check, in one document. `report()` runs a whole battery of modules and writes the result to an HTML file you can open in a browser:

```{r}
#| eval: false
report(paper, output_file = "good_paper_report.html")
```

This writes `good_paper_report.html` into your working directory. Open it, and you will see the same traffic lights and summaries from Step 2, but now for every module the report includes — preregistration, funding, conflicts of interest, power, statistical reporting, and more — all in one place, ready to read or share.

You already have the bad paper read in from Step 2 as `bad_paper`, so running the same check on it just means replacing `paper` with `bad_paper`:

```{r}
#| eval: false
report(bad_paper, output_file = "bad_paper_report.html")
```

This produces the same kind of report, but with several more red and yellow lights: a missing ethics statement, a missing funding statement, an unwarranted causal claim, p-values that do not match their own reported statistics, and more. Nothing about the *process* changes between the two papers — only what the report finds.

## Step 4: Check a shared repository

So far we have checked what a paper *says*. Metacheck can also check what a paper's authors actually *shared* — the data and code repository itself — independently of any manuscript. `report_repository()` is the function for this: give it a folder, and it runs every repository-oriented check (file organisation, the code, the data, and the codebook) and writes a report, the same way `report()` did in Step 3.

Download the bad paper's repository as a zip file (the same "Download ZIP" button you would click on GitHub) and unpack it:

```{r}
#| eval: false
download.file(
  "https://github.com/Lakens/stroop_effect_paper_bad/archive/refs/heads/main.zip",
  destfile = "bad_repo.zip", mode = "wb"
)
unzip("bad_repo.zip", exdir = "bad_repo")
```

Then point `report_repository()` at the unpacked folder — there is no manuscript involved in this step, so there is no paper object to read in first:

```{r}
#| eval: false
report_repository("bad_repo/stroop_effect_paper_bad-main",
                  output_file = "bad_repo_report.html")
```

Open `bad_repo_report.html`, and you will see a **yellow** light on the repository's overall organisation: among other things, Metacheck cannot confidently classify every file it finds, and flags a file naming problem. Try the same two lines on the good paper's repository (<https://github.com/Lakens/stroop_effect_paper>) and compare the result — it is organised into clearly labelled `data/`, `code/`, `output/`, and `documentation/` folders, with a complete codebook, which is exactly what this check is designed to reward.

## What you have just done

In four short steps, you:

1. Read a PDF into R as a paper object (`convert()` + `read()`).
2. Ran a single check on it (`module_run()`).
3. Generated a full, shareable report from a paper (`report()`).
4. Generated a full, shareable report from a shared repository, with no manuscript involved (`report_repository()`).

These four functions cover the large majority of what you will do with Metacheck day to day. Everything else in this book is about the details: which modules exist and what each one checks ([Modules](mod-power.qmd)), how reading in a paper actually works ([Reading in a Paper](reading-a-paper.qmd)), how to search a paper's text yourself ([Text Search](text-search.qmd)), and how to turn on a large language model for the modules that can use one ([Using Large Language Models](llms.qmd)).
