---
title: "Function reference for new users"
df-print: paged
vignette: >
  %\VignetteIndexEntry{Function reference for new users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
```

```{r}
#| include: false
library(metacheck)
```

This article lists the functions in Metacheck with a short, plain-language
explanation of what each one does and a small example. It is meant for people
who are starting to use the package and want a map of what is available.

It is organised in two parts:

1. **Public functions** — the functions you call directly, grouped by task.
2. **Internal functions** — helpers used inside the package, listed for
   completeness.

Examples that work offline are run when this page is built, so their output is
real. Examples that need the internet (looking up DOIs, downloading from
repositories, querying an LLM) are shown but not run; they are marked with a
comment.

Two example data sets ship with the package and are used throughout:

- `demopaper()` is a single small paper object, good for quick examples.
- `psychsci` is a list of 250 open-access papers from *Psychological Science*.

```{r}
paper <- demopaper()
```

# Public functions

## Getting a paper to work with

Everything in Metacheck operates on a *paper object* (or a list of them). These
functions create or load one.

**`demopaper()`** — returns a ready-made example paper, so you can try things
without loading a file.

```{r}
paper <- demopaper()
class(paper)
```

**`test_paper(text)`** — builds a minimal paper from a string of text. Handy for
testing a check on a sentence you type yourself.

```{r}
tp <- test_paper("There was an effect, t(28) = 2.40, p = .02, d = 0.45.")
tp$text$text
```

**`demofile(ext)`** — returns the path to a bundled example file of a given type
(for trying the import functions).

```{r}
demofile("pdf")
```

**`read(file_path)`** — reads a paper from a Metacheck JSON file (or a folder of
them) into a paper object.

```{r}
#| eval: false
paper <- read("mypaper.json")
```

**`paper(id, ...)`** — constructs a paper object from components directly.
**`paperlist(...)`** — combines several papers into one list that the checks can
loop over.

```{r}
pl <- paperlist(demopaper(), test_paper("t(10) = 2.0, d = 0.6."))
length(pl)
```

**`paper_id(paper)`** — the paper's identifier.
**`paper_table(paper, table)`** — pull one of the paper's component tables
(e.g. `"info"`, `"author"`, `"bib"`).
**`ref_table(paper)`** — the reference list as a tidy table.

```{r}
paper_id(paper)
ref_table(paper)
```

**`paper_validate(paper)`** — checks a paper object against the expected schema.
**`paper_write(paper, file_name)`** — saves a paper object to a JSON file.
**`fig_image_view(paper, figure_id)`** — displays a figure image from the paper.

```{r}
#| eval: false
paper_validate(paper)
paper_write(paper, "mypaper.json")
```

## Importing papers from PDFs and XML

These convert source documents (usually a PDF) into the Metacheck JSON format.
Most need an external service (a GROBID server or the bibr backend), so they are
shown but not run here.

**`convert(file_path, save_path)`** — the main entry point: convert a document to
Metacheck JSON, picking a method automatically.
**`convert_grobid(...)`** / **`convert_bibr(...)`** — convert using a specific
backend.
**`grobid_to_bibr(xml_path)`** — turn GROBID TEI XML into the bibr structure.
**`format_bib_authors(authors)`** — format an author list for a bibliography.

```{r}
#| eval: false
# needs a running GROBID server or the bibr backend
paper <- convert("mypaper.pdf", save_path = "out/")
```

## Searching the text

**`search_text(paper, pattern)`** (also available as **`text_search()`** — they
are the same function) — find text matching a regular expression.

```{r}
search_text(paper, "significant") |> head(3)
```

Use `return = "match"` to get just the matched pieces, `fixed = TRUE` to match
literally, and `exclude = TRUE` to get the rows that *don't* match.

```{r}
search_text(paper, "significant", return = "match")
search_text(paper, "p = 0.005", fixed = TRUE)
nrow(search_text(paper, "significant", exclude = TRUE))
```

**`expand_text(results, paper, expand_to)`** (also **`text_expand()`**) — add a
column with surrounding context (e.g. the whole sentence or paragraph) to a
search/extraction result.

```{r}
pv <- extract_p_values(paper)
expand_text(pv, paper, expand_to = "sentence")$expanded[1]
```

## Extracting statistics and links

**`extract_eq(paper)`** — list every "name `<comparator>` value" equation in the
text (test statistics, means, effect sizes, …). This is the canonical place to
get statistics; checks should read from here rather than re-scanning the text.

```{r}
extract_eq(paper) |> head()
```

**`extract_p_values(paper)`** — the p-values, with comparator and numeric value.

```{r}
extract_p_values(paper)[, c("text", "p_comp", "p_value")]
```

**`extract_urls(paper)`** — text rows containing a URL or DOI.

```{r}
extract_urls(paper)[, c("text_id", "text")] |> head(3)
```

**`stats(text)`** — run statcheck-style consistency checks on reported
statistics.
**`json_expand(table, col)`** — expand a column of JSON strings into columns.

```{r}
#| eval: false
stats(paper)
```

## Running checks (modules)

Checks in Metacheck are *modules*. These functions list, describe, and run them.

**`module_list()`** — the available modules.

```{r}
module_list()[, c("name", "title", "section")] |> head()
```

**`module_run(paper, module)`** — run a module; returns a list with `table`,
`summary_table`, `traffic_light`, `summary_text`, and `report`.

```{r}
out <- module_run(paper, "all_urls")
out$traffic_light
```

**`module_help(module)`** / **`module_info(module)`** — documentation and
metadata for a module.
**`module_template(module_name)`** — scaffold a new module file.
**`get_prev_outputs(module, item)`** — inside a module, read the output of a
module that ran earlier.

```{r}
module_info("all_urls")$title
```

## Building a report

**`report(paper, modules)`** — run a set of modules and render a full report
document.
**`report_module_run(...)`** — run the modules that make up a report.
**`module_report(module_output)`** — turn one module's output into its report
section.
**`report_qmd(...)`** — produce the Quarto source for a module's section.

```{r}
report_txt <- module_report(out)
class(report_txt)
```

```{r}
#| eval: false
report(paper, modules = c("all_urls", "stat_effect_size"),
       output_file = "report.html")
```

## Report-building helpers

Small helpers modules use to format their report sections.

**`scroll_table(table)`** / **`report_table(table)`** — render a data frame as an
HTML table.
**`collapse_section(text, title)`** — wrap text in a collapsible block.
**`link(url, text)`** — make an HTML link.
**`plural(n)`** — `"s"` when `n != 1`, else `""` (for grammatical messages).
**`format_ref(bib)`** — format a citation from a bibentry.

```{r}
plural(1)
plural(2)
link("https://example.org", "example")
format_ref(citation("metacheck"))
```

## Looking up DOIs and metadata

**`doi_clean(doi)`** — normalise a DOI string.
**`doi_valid_format(doi)`** — does it look like a DOI?
**`doi_resolves(doi)`** / **`doi_lookup(doi)`** — check/look up a DOI online.

```{r}
doi_clean("https://doi.org/10.1234/ABC")
doi_valid_format("10.1234/abc")
```

```{r}
#| eval: false
doi_lookup("10.1234/abc")   # needs internet
```

**`crossref_doi(doi)`** / **`crossref_query(ref)`** — Crossref metadata lookup.
**`openalex_doi(doi)`** / **`openalex_query(title)`** — OpenAlex lookup.
**`datacite_doi(doi)`** — DataCite lookup.
**`add_bib_match(paper)`** — match a paper's references to online records.

```{r}
#| eval: false
crossref_doi("10.1177/0956797614520714")  # needs internet
```

## Databases of comments, replications, retractions

**`pubpeer_comments(doi)`** — PubPeer comments for a DOI.
**`FLoRA()`** / **`FLoRA_update()`** / **`FLoRA_date()`** — the FLoRA replication
database (loaded, refreshed, and dated).
**`retractionwatch()`** (alias **`rw()`**) / **`rw_update()`** / **`rw_date()`** —
the Retraction Watch database.

```{r}
#| eval: false
pubpeer_comments("10.1177/0956797614520714")  # needs internet
rw()                                            # downloads the database
```

## Finding and downloading shared files

Each repository has a `*_links()` function that finds links in a paper, an
`*_info()` function that fetches metadata, and (where supported) a
`*_file_download()` function.

**OSF:** `osf_links()`, `osf_info()`, `osf_file_download()`, `osf_type()`,
`osf_check_id()`, `osf_api_check()`, `osf_delay()`, `osf_get_all_pages()`,
`osf_preprint_list()`.

**GitHub:** `github_links()`, `github_info()`, `github_repo()`,
`github_files()`, `github_languages()`, `github_readme()`.

**Zenodo:** `zenodo_links()`, `zenodo_info()`, `zenodo_file_download()`.

**ResearchBox:** `rbox_links()`, `rbox_info()`, `rbox_file_download()`.

**AsPredicted:** `aspredicted_links()`, `aspredicted_info()`.

**Local files:** `local_files(path)` — list files in a local folder.

The `*_links()` functions work offline on an already-loaded paper:

```{r}
osf_links(paper)[, c("href", "text_id")]
```

The `*_info()` and download functions need the internet:

```{r}
#| eval: false
osf_info("https://osf.io/5tbm9")          # fetch metadata
osf_file_download("5tbm9", "downloads/")  # download files
github_readme("scienceverse/metacheck")
```

## Asking a large language model

**`llm(text, ...)`** — send text to a language model and get a structured
answer.
**`llm_use(...)`** — set which model/platform to use.
**`llm_model(model)`** / **`llm_model_list(platform)`** — pick or list models.
**`llm_max_calls(n)`** — cap how many calls a run may make.

```{r}
#| eval: false
llm_use("ollama")                       # choose a backend
llm("Summarise this abstract: ...")     # query the model
```

## Checking shared code

**`code_read(file_path)`** — read a code file.
**`code_parse_r(file_path)`** — parse R code into a table of calls.
**`code_remove_comments(code_text, lang)`** — strip comments from code.

```{r}
code_remove_comments("x <- 1 # set x", lang = "R")
```

## ORCID and contributor info

**`check_orcid(orcid)`** — validate an ORCID and fetch the person.
**`get_orcid(family, given)`** — look up an ORCID by name.
**`orcid_person(orcid)`** — a person's ORCID record.
**`credit_roles()`** — the CRediT contributor-role taxonomy.

```{r}
credit_roles() |> head()
```

```{r}
#| eval: false
check_orcid("0000-0002-0247-239X")   # needs internet
```

## File types

**`filetype(filename)`** — a coarse type for a filename (e.g. data, code,
document).
**`file_category(contents)`** — categorise file contents.

```{r}
filetype("analysis.R")
filetype("data.csv")
```

## Validation utilities

**`validate(gt, module)`** — compare a module's current output against a coded
ground-truth file.
**`accuracy(expected, observed)`** — accuracy measures (sensitivity,
specificity, d-prime, …) from two logical vectors.

```{r}
accuracy(c(TRUE, TRUE, FALSE, FALSE), c(TRUE, FALSE, FALSE, FALSE))
```

## General utilities

**`email(email)`** — set the contact email used for polite API access.
**`online(url)`** — is a URL reachable?
**`verbose(verbose)`** — get/set whether functions print progress.
**`pb(total)`** — a progress bar.
**`rep_if(x, y, replace)`** — replace values equal to `y` with `replace`.
**`path_sanitize(path)`** — make a string safe to use as a file path.
**`message(...)`** — Metacheck's message wrapper.
**`logger()`, `logpath()`, `lastlog()`** — write and read run logs.

```{r}
rep_if(c(1, NA, 3), NA, 0)
path_sanitize("results: study/1?.csv")
```

```{r}
#| eval: false
email("you@example.org")
online("https://osf.io")
```

# Internal functions

These are not part of the public interface and may change without notice. They
are listed so the documentation is complete. Several are exported only for
testing and begin with a dot.

```{r}
#| echo: false
ns <- getNamespace("metacheck")
is_fun <- function(n) is.function(get(n, envir = ns))
exported <- getNamespaceExports("metacheck")
all_fun <- Filter(is_fun, ls(ns, all.names = TRUE))
internal <- sort(setdiff(all_fun, exported[!grepl("^[.]|^%", exported)]))
internal <- internal[!grepl("^%", internal)]
internal
```

```{r}
#| echo: false
cat(length(internal), "internal functions.\n")
```
