---
title: "Data Check"
vignette: >
  %\VignetteIndexEntry{Data Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

The `data_check` module is the foundation for all of Metacheck's data-oriented checks. It takes the files shared with a paper — from a linked repository or a local folder — and does three things:

1. **Classifies every file** by what it is: tabular *data*, *code*, a *codebook*, a *README*, an *archive*, or a raw file (image, PDF, binary).
2. **Reads each tabular data file** and describes every column along several independent **facets** — how it is stored, its measurement level, what it measures, and how it functions — together with basic **summary statistics**.
3. **Makes the extracted columns available** to the other data modules — `data_validate`, `excel_check`, `codebook_check`, and `psychds_check` all build on `data_check`'s output rather than re-reading the files.

On its own, `data_check` gives you a machine-readable inventory of what was shared and what each data column contains. It is also the first module in the data pipeline: run it once, and the downstream modules reuse its results.

::: callout-note
This chapter runs entirely offline. We build a small example repository in a temporary folder and point `data_check` at it with `local_path`. The same calls work against a real downloaded repository — see the [Local Files](local-files.qmd) chapter.
:::

## A worked example

We create a tiny study folder with one data file and a code file:

```{r}
repo <- file.path(tempdir(), "data_check_demo")
dir.create(file.path(repo, "data"), recursive = TRUE, showWarnings = FALSE)

study <- data.frame(
  participant_id = 1:20,
  condition      = rep(c("control", "treatment"), each = 10),
  age            = sample(18:65, 20, replace = TRUE),
  rt             = round(runif(20, 300, 900)),
  score          = round(rnorm(20, 50, 10), 1),
  passed         = sample(c(0, 1), 20, replace = TRUE)
)
utils::write.csv(study, file.path(repo, "data", "study.csv"), row.names = FALSE)
writeLines("# analysis\nd <- read.csv('study.csv')",
           file.path(repo, "analysis.R"))
```

Now run the module. `test_paper()` supplies a minimal paper object with no repository links, and `local_only = TRUE` means only the local folder is inspected.

```{r}
mo <- module_run(test_paper(), "data_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## File classification

The first thing `data_check` produces is a classification of every file it found. This is available in the `structure` element:

```{r}
mo$structure[, c("file_name", "file_type", "data_type")] |>
  knitr::kable()
```

Each file is sorted into a category by its name and extension:

| Category | Examples |
|---|---|
| `data` | `.csv`, `.tsv`, `.xlsx`, `.sav`, `.dta`, `.rds` |
| `code` | `.R`, `.Rmd`, `.qmd`, `.py`, `.do`, `.sps` |
| `codebook` | files whose name or contents look like a data dictionary |
| `readme` | `README`, `read me` |
| `archive` | `.zip`, `.tar`, `.7z` |
| `raw` | images, PDFs, and other non-tabular binaries |

## The data tree

To make the layout of a repository easy to see at a glance, `data_check`'s report includes a **data tree** — the folder structure of the repository drawn as an indented tree. Directories are listed before files and both are sorted alphabetically, so the same repository always produces the same tree. In the rendered report the tree appears inside a collapsible *"Data Tree"* callout (one per repository, when several are linked).

The tree is part of the module's `report`. We can pull the tree lines out of the report to show them here:

```{r}
tree_lines <- grep("^[│├└ ]", unlist(strsplit(mo$report, "\n")), value = TRUE)
cat(tree_lines, sep = "\n")
```

For our example folder — a `data/` directory holding `study.csv`, plus a top-level `analysis.R` — the tree makes the organisation obvious. On a real repository this is often the quickest way to spot problems: data scattered across many ad-hoc folders, deeply nested paths, or files sitting loose at the top level.

This same tree is the starting point for the [Psych-DS Check](mod-psychds-check.qmd), which compares a repository's actual tree against the tree the Psych-DS standard says it *should* have.

## Column facets

For every readable tabular file, `data_check` reads the data and describes each column. Earlier versions gave each column a single `col_type` value — but that conflated things that are not really alternatives: *how* a value is stored (a number), *what level* it is measured on (nominal, ordinal, …), and *what it actually measures* (a reaction time, an age). Following the [DDI](https://ddialliance.org/) metadata standard, `data_check` now describes each column with several **orthogonal facets** instead, so these independent properties are recorded independently. The per-column results are in the `table` element:

```{r}
mo$table[, c("column_name", "representation", "measurement_level",
             "concept", "role", "unit")] |>
  knitr::kable()
```

The facets are:

| Facet | Values | What it captures (DDI analogue) |
|---|---|---|
| `representation` | `numeric`, `text`, `datetime`, `code`, `empty` | how the value is stored (RepresentedVariable representation) |
| `measurement_level` | `nominal`, `ordinal`, `interval`, `ratio` | the Stevens level (DDI `@classificationLevel`) |
| `concept` | `reaction_time`, `accuracy`, `age`, `gender`, `race`, `likert`, `condition`, `id`, `date`, `timestamp`, … | what the column *measures* (DDI Variable → Concept) |
| `role` | `identifier`, `measure`, `condition`, `timestamp` | how it functions in the dataset (DDI VariableRole) |
| `unit` | `seconds`, `milliseconds`, `years`, … | the unit of measure (DDI UnitType) |
| `quality` | `ok`, `empty`, `constant` | a data-quality *state*, kept separate from type |

Because the facets are independent, an `age` column is `numeric` **and** `ratio` **and** the `age` concept **and** measured in `years` — each fact recorded in its own field rather than crammed into one label. A `participant_id` is `text` with the `identifier` role and the `id` concept; a Likert item is `numeric`, `ordinal`, with the `likert` concept.

Concepts are detected by **name-and-value agreement**: a column is only tagged (say) `reaction_time` when its name looks like a reaction time *and* its values are plausible durations, which keeps false positives low. The demographic concepts (`age`, `gender`, `race`) use the exported helper `data_check_demographic()`; a reaction-time column additionally seeds its `unit` (`milliseconds` vs `seconds`) from the value magnitude.

`data_check` keeps this rule-based path fully offline and deterministic. When you enable an LLM (see below), it fills in the `concept` and `measurement_level` for columns the rules left blank — for example a reaction time hidden behind a cryptic name like `q3`.

::: callout-note
The old single `col_type` field has been replaced by these facets. If you have code that read `mo$table$col_type`, read the relevant facet instead — most often `representation` (numeric vs text) or `concept`.
:::

## Unit of observation

`data_check` also infers, for each data file, what **one row represents** — a `person`, a `trial`, a `session`, or a `dyad` (DDI's `analysisUnit`). This matters because it tells a reviewer whether rows are independent (one per participant) or nested (many trials per participant), which changes how the data should be analysed. It is inferred from the identifier column(s) and whether they are unique per row: a unique-per-row id is person-level; a repeating id alongside a `trial`/`item` column is trial-level (long format); a repeating id with a `session`/`wave` column is a repeated-measures unit; two id columns suggest a dyad.

```{r}
unique(mo$table[, c("source_file", "analysis_unit")]) |>
  knitr::kable()
```

Each row of our `study.csv` is one participant, so the file is `person`-level. When a repository **mixes** units — say a person-level file next to a trial-level file — the report flags it, because combining files at different levels without aggregating is a common and easily-missed error.

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `"green"` when data files were read successfully, `"na"` when there is nothing to check |
| `$summary_text` | plain-text summary: how many files, how many columns |
| `$summary_table` | per-paper counts |
| `$table` | one row per column, with the facets (`representation`, `measurement_level`, `concept`, `role`, `unit`, `quality`), the inferred `analysis_unit`, and summary statistics |
| `$structure` | one row per file, with `file_type` and `data_type` |
| `$previews` | the full data frames that were read (used by `data_validate`) |

The `$structure` and `$previews` elements are what the downstream modules consume, which is why you rarely need to run `data_check` by itself — running `data_validate` or `codebook_check` triggers it automatically and reuses the result.

## Downloading files

When a paper links to an online repository, `data_check` downloads the files it needs into a shared on-disk cache (reused across runs, so a file is never fetched twice). The `download` argument controls **what** is fetched:

- `download = "data"` (the default) fetches only the machine-readable files the checks analyse — tabular data plus codebook/README files.
- `download = "all"` fetches **every** file in the repository (code, materials, PDFs, assets, ...). This is the right choice when building a complete data archive with `convert_psychds()`.
- `download = "none"` (or `FALSE`) downloads nothing — files are only classified by name.

```{r}
#| eval: false
# default: fetch only the readable data + codebook/README files
module_run(paper, "data_check")

# fetch everything, for building a full data archive
module_run(paper, "data_check", download = "all")
```

### Size caps

Two caps bound how much is downloaded per repository. They work differently:

- **`max_file_size`** (default **100 MB**) is a **per-file filter**: a single file larger than this is skipped on its own, and the rest of the repository still downloads.
- **`max_download_size`** (default **500 MB**) is a **per-repository gate**: if the total of the files that would download exceeds it, the *whole* repository is skipped.

Set either to `Inf` for no cap. A file whose size cannot be determined at all (no size in the listing and no `Content-Length` header) also causes its repository to be skipped, so nothing huge is ever fetched blind. Every skip is reported inline and recorded, so you can raise the relevant cap and re-run.

```{r}
#| eval: false
module_run(paper, "data_check", download = "all",
           max_file_size = 200, max_download_size = 2000)
```

### Skipping asset types, and peeking inside zips

When building an archive you often want the *data* but not the stimuli or media (images, audio, video) — those are usually better linked to than mirrored. Two options handle this:

- **`skip_types`** — a vector of file types never to download, even under `download = "all"`. Most usefully `skip_types = "asset"` leaves out stimuli/media. Skipped files are still listed in the manifest with the reason, so the release can point to them.
- **`peek_zips = TRUE`** — before downloading a `.zip`, look inside it via an HTTP range request (fetching only the zip's tail, not the whole file) and only download zips that actually contain data or a codebook. A zip of only stimuli is left to be linked. Downloaded data-zips are unpacked and their inner data files added to the archive (inner assets dropped). See `zip_peek()` and `zip_decision()`.

```{r}
#| eval: false
# archive-building recipe: everything except media, peeking inside zips
module_run(paper, "data_check", download = "all",
           skip_types = "asset", peek_zips = TRUE)
```

### A file manifest

Pass `manifest = "some/dir"` to write a per-paper JSON manifest listing every repository file with its URL, size, type, and whether it was downloaded (and if not, why). Unsized files are resolved with a lightweight `HEAD` request so the manifest carries real sizes — useful for auditing a corpus or choosing size caps before a large archive build.

```{r}
#| eval: false
module_run(paper, "data_check", download = "none", manifest = "manifests")
```

## Using an LLM

With `llm_use(TRUE)`, `data_check` adds things the rules cannot do reliably:

- **Concepts for cryptically-named columns** — the rules only tag a concept when the name gives it away, so a reaction time called `q3` stays blank; the LLM fills it in.
- **Refined measurement levels** for ambiguous integer columns (nominal vs ordinal vs ratio).
- **A study `group`** for each file, inferring which experiment or study a data file belongs to from its name and context — useful when a repository bundles several studies together.

```{r}
#| eval: false
llm_use(TRUE)
llm_model("groq/openai/gpt-oss-20b")
mo <- module_run(test_paper(), "data_check",
                 local_path = repo, local_only = TRUE)
mo$table$concept   # filled in for cryptic names
mo$table$group     # study grouping, otherwise NA
```

See the [LLMs](llms.qmd) chapter for how to configure a model.

## Related

- [Data Validate](mod-data-validate.qmd) runs automated data-quality checks on the columns `data_check` extracts.
- [Excel Check](mod-excel-check.qmd) inspects `.xlsx` files for non-machine-readable formatting.
- [Codebook Check](mod-codebook-check.qmd) checks whether the columns are documented.
- [Psych-DS Check](mod-psychds-check.qmd) checks the repository against the Psych-DS standard.
