---
title: "Repository Check"
vignette: >
  %\VignetteIndexEntry{Repository Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `repo_check` module finds links to **research repositories** in a paper — the Open Science Framework (OSF), GitHub, ResearchBox, PsychArchives, and Zenodo — retrieves the list of files in each, and reports what is shared. It is the foundation for every other module that reads repository files: [code_check](mod-code-check.qmd) and [data_check](mod-data-check.qmd) both build on the file list `repo_check` produces.

Beyond listing files, `repo_check` runs a set of checks on what it finds:

- whether each repository has a **README**;
- whether files are bundled into **archives** rather than shared individually;
- whether any files are **unclassifiable** by name or extension;
- whether **file names** follow machine-parseable conventions;
- whether the studies the manuscript names match how the repository's files are **grouped**;
- whether proprietary **E-Prime** binary files (`.edat`/`.emrg`) have a matching plain-text export;
- whether any files sit behind **restricted or embargoed access**; and
- whether an OSF **registration** points at a closed or inaccessible source project.

::: callout-note
This module makes **live network calls** to the linked repositories. You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` links to several repositories.

```{r}
paper <- demopaper()
mo <- module_run(paper, "repo_check")
mo$traffic_light
cat(mo$summary_text)
```

The `table` lists every file found across all linked repositories:

```{r}
mo$table[, c("repo_name", "file_name", "file_type", "file_size")] |>
  head(10) |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "repo_check")
mo$summary_table
```

## Repositories it searches

`repo_check` looks for links to five repository platforms, each via its own link-finding helper (`osf_links()`, `github_links()`, `rbox_links()`, `psycharchives_links()`, `zenodo_links()`) and its own file-listing call:

| Platform | Link finder | How files are listed |
|---|---|---|
| OSF | `osf_links()` | recursive `osf_info()`, including registrations (see below) |
| GitHub | `github_links()` | `github_tree_files()` — one repo-metadata + one full-tree API call, listing the repository in full |
| ResearchBox | `rbox_links()` | `rbox_file_download()` (downloads and unzips the box) |
| PsychArchives | `psycharchives_links()` | `psycharchives_file_download()` (DSpace REST API; lists without downloading) |
| Zenodo | `zenodo_links()` | `zenodo_info()` |

### OSF registrations follow their source project

An OSF **registration** is a locked snapshot. Many "Open-Ended Registration" entries (common in older schemas) only *describe* what was shared, with the real files living in the project the registration was made from (`registered_from`). Since the manuscript usually links only the registration, `repo_check` follows `registered_from` automatically:

- if the source project is **public**, its files are listed under the source project's own URL (not the registration's);
- if the source project is **closed or inaccessible**, the report flags this explicitly, distinguishing two cases: the registration itself still holds a mirrored copy of the files (nothing is lost, but the permanent home is unreachable), or no copy exists anywhere (the files are genuinely unreachable).

### GitHub repositories are listed in full

Every GitHub repository is listed in full, whatever its size, exactly as OSF, Zenodo, ResearchBox, PsychArchives and local directories have always been treated. Listing a repository is cheap — it costs two API calls and transfers no file contents — so there is nothing to configure and no size threshold to set.

There remain two reasons a GitHub repository can fail to be listed, and neither is one a setting can fix. Either the repository is invalid or inaccessible (it has been renamed, deleted, or made private), or it exceeds GitHub's own limit of 100,000 items per tree request, in which case GitHub returns a listing it marks as truncated and metacheck reports it as too large to list rather than silently using a partial one. Repositories in either state are recorded in `$gated_repos` (see below) with the reason, so a report can explain *why* a paper produced no GitHub files instead of simply reporting none found.

Note that this concerns *listing* only. How much of a repository is ever **downloaded** is still capped, by the per-file and per-repository size limits in `data_check` described in the [data_check](mod-data-check.qmd) chapter.

::: {.callout-note}
## A change from earlier versions

Earlier versions of `repo_check` took three arguments — `github_gate`, `github_max_repo_size_mb` and `github_max_files` — that skipped listing a repository over a size threshold. These have been removed, so code passing them will now fail. Delete them from any script that used them; no replacement is needed, because listing is no longer gated.
:::

## Study grouping

`repo_check` runs a **preliminary** classification and grouping pass over every file it finds:

- `data_classify_files()` labels each file's `data_type` (data, code, readme, archive, documentation, output, …) from its name/path/extension;
- `.data_doc_role()` further labels documentation-type files (`readme`, `license`, `codebook`, …);
- `data_group_llm()` assigns each file to a study `group`, using deterministic path/repo-split rules first and falling back to an LLM (`llm_use(TRUE)`) only when those are ambiguous — see the [LLMs](llms.qmd) chapter. Root-level files that are collection-wide by nature (a top-level README, `ro-crate-metadata.json`, or LICENSE) are excluded from grouping, since they don't belong to any one study.

This pass is **preliminary**: `repo_check` never downloads file contents (except where a platform's own listing call requires it, like ResearchBox), so it classifies from names and paths only. [data_check](mod-data-check.qmd) reclassifies afterwards using file *contents* too, and its classification is authoritative for where a file is placed in a paper's structure. `repo_check`'s own pass exists to power its own report — the file-naming check, the classification dropdown, and the study/repository mismatch check below.

### Study/repository mismatch

If the manuscript explicitly names studies (e.g. "Study 1", "Study 2"), `repo_check` compares that roster against the studies the *files* actually separate into. A mismatch — a named study with no matching files, or a file-derived group not named in the manuscript — is flagged in the report. A manuscript that never names studies explicitly has no roster to compare against, so this check is silently skipped rather than treated as a mismatch.

## File naming conventions

`check_file_naming()` checks every file name for machine-parseability: no spaces or special characters, and every file classifiable by name or extension (`data_type != "unknown"`). Issues are split into two severities:

- **bad** — a real naming problem (should be fixed);
- **suggestion** — not required, but a good habit (e.g. a more descriptive name).

```{r}
mo$naming_issues |> head(5) |> knitr::kable()
```

## Special file types

### Proprietary E-Prime files

E-Prime's `.edat`/`.edat2` (and `.emrg`/`.emrg2` merge files) are proprietary binary formats that only the E-Prime software can open — metacheck neither downloads nor reads them. The analysable data lives in E-Prime's plain-text export instead. `repo_check` flags any `.edat`/`.emrg` file that has **no matching `.txt` export** alongside it (matched by file stem), so authors know to also share the readable export (File → Export in E-Prime).

### Restricted-access files

A PsychArchives item flagged `restrictedAccess`/`embargoedAccess` hides its files behind institutional (SSO) login — they cannot be fetched programmatically. `repo_check` flags any repository with restricted files, since metacheck cannot examine them and other researchers cannot reuse them without requesting access.

### Archives, and looking inside a `.zip`

An archive shared as a single file hides what is in it. By default `repo_check` therefore reads each `.zip`'s own listing and reports the files inside it, in place of the archive as one opaque row. This costs one HTTP range request per archive and downloads none of its contents: a `.zip` stores a directory of what it holds at the end of the file, and a server that honours range requests will return just that stretch of bytes. The helper that does this is `zip_peek()`, and you can call it yourself on any archive URL.

Turn it off to leave archives as single rows and make no per-archive request at all:

```{r}
#| eval: false
module_run(paper, "repo_check", peek_zips = FALSE)
```

This works for `.zip` only. `.7z`, `.rar` and `.tar.gz` keep no equivalent directory that can be read from the end of the file, so they would have to be downloaded whole, and some formats metacheck cannot read at all. `repo_check` warns when an archive is in one of these formats, because that limits how discoverable and reusable its contents are. An archive whose listing cannot be read for any other reason — a server that ignores range requests, for instance — is simply left as a single row, and `data_check` opens it after downloading, as before.

## A clean example and one with problems

The demo paper links to three repositories, and running `repo_check` on it produces a **yellow** light — there are problems to address. The summary text reports exactly what:

```{r}
mo$traffic_light
cat(mo$summary_text)
```

Here the issues are a **missing README** and an **archive file** whose contents could not be examined. The per-paper summary table counts each file type — note `files_readme` is 0 (no README was shared) and `files_zip` is 1 (one archive):

```{r}
mo$summary_table |>
  knitr::kable()
```

We can see the specific archive file that was flagged in the full file table:

```{r}
mo$table[mo$table$file_type %in% "archive", c("repo_name", "file_name")] |>
  knitr::kable()
```

A **clean** result, by contrast, would include a README in every linked repository (`files_readme` ≥ 1 for each), share files individually rather than as a zip (`files_zip` of 0), have no unclassifiable or badly-named files, agree with the manuscript's study roster, and have no closed registrations — producing a green light with no items to address.

## Options

`repo_check` accepts arguments for working with **local files** instead of (or in addition to) online repositories:

```{r}
#| eval: false
# also include files from a local folder
module_run(paper, "repo_check", local_path = "path/to/downloaded/files")

# only check a local folder, skip all online lookups
module_run(paper, "repo_check", local_path = "path/to/files", local_only = TRUE)
```

`local_only = TRUE` is useful when you cannot or do not want to contact external services — for example when checking a reviewer submission you downloaded as a zip. See the [Local Files](local-files.qmd) chapter for details.

`model` and `params` are passed through to `data_group_llm()` for study grouping, and only take effect when `llm_use(TRUE)` and the deterministic grouping rules cannot place a file:

```{r}
#| eval: false
module_run(paper, "repo_check", model = "gpt-4o-mini", params = list(temperature = 0))
```

## The traffic light and what you get back

| Light | Meaning |
|---|---|
| **green** | repositories were found; every one has a README; no archive files, unclassifiable files, or bad file names; no study/repository mismatch; and no closed OSF registrations |
| **yellow** | at least one repository is missing a README, an archive file was found, files could not be classified, real naming problems were found, the manuscript's study roster disagrees with the repository's file grouping, or a registration points at a closed source project |
| **na** | no repository links were found in the paper at all |

| Element | What it contains |
|---|---|
| `$table` | one row per file across all repositories: `repo_url`, `file_name`, `file_path`, `file_url`, `file_location`, `file_size`, `file_type`, `data_type`, `doc_role`, `group`, `paper_id` |
| `$summary_table` | per-paper aggregates: `repo_n`, `files_n`, `files_data`, `files_code`, `files_readme`, `files_zip`, `files_unknown`, `naming_issues`, `roster_mismatch` |
| `$naming_issues` | full output of `check_file_naming()` — one row per issue, with severity |
| `$roster_check` | the manuscript-vs-file study roster comparison (`roster`, `missing`, `extra`), or `NULL` if not applicable |
| `$gated_repos` | repositories found but not listed (e.g. a GitHub repo over the size gate, or a private OSF component), with the reason |
| `$group_no_evidence` | `TRUE` if study grouping had no evidence to work from |
| `$report` | formatted report sections: repositories, files, README, archives, E-Prime, restricted access, closed registrations, unclassified files, roster mismatch, naming, and a full classification breakdown |
| `$summary_text` | plain-text bulleted summary of each check |

## Related

The companion [code_check](mod-code-check.qmd) module analyses the actual code files that `repo_check` discovers. [data_check](mod-data-check.qmd) re-downloads and reclassifies the data files using their contents (not just names and paths), and runs the [data-quality checks](mod-data-validate.qmd) on the columns it finds.
