---
title: "Replication Check"
vignette: >
  %\VignetteIndexEntry{Replication Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## The basics

When a paper cites an earlier ("original") study, it is useful to know whether anyone has since tried to redo that study to see if the same result shows up again — this is called a replication. `ref_replication` looks at the studies a paper cites and checks whether any of them have a known replication attempt on record, using an online database of replications called [FLoRA](https://forrt.org/).

Here is the module running on a demo paper, which cites an original study that has a recorded replication. Because this module needs an internet connection to look up the database, running it can take a moment:

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "ref_replication")
```

Running a module always works the same way: `module_run(paper, "module_name")`. The printed result above is a short report: a coloured light (called a traffic light, after the red/yellow/green of a real traffic light) and a one-line summary.

If you want to work with the result instead of just reading it, save it to a variable and look inside it:

```{r}
mo <- module_run(paper, "ref_replication")
mo$table[, c("doi", "replication_doi", "replication_url")] |>
  knitr::kable()
```

`mo$table` matches each cited original study to its replication(s), identified by their DOI (a permanent identifier for a published paper). A match only means a replication exists in the database — it does not tell you whether that replication succeeded or failed; for that you need the `show_outcomes` option described later in this chapter. Either way, treat a match as an invitation to go look at the replication yourself, not as a final verdict on the original study.

## What it checks

The `ref_replication` module cross-references the paper's **cited references** against the [FLoRA](https://forrt.org/) replication database. If you cite an original study that has since been the subject of a replication attempt, that is highly relevant context — especially if the replication failed. The module flags cited originals that have known replications, and can report the replication's outcome.

::: callout-note
This module makes **live network calls** to the FLoRA replication database. You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` cites an original study that has a recorded replication.

```{r}
paper <- demopaper()
mo <- module_run(paper, "ref_replication")
mo$traffic_light
mo$summary_text
```

The `table` matches each cited original to its replication(s):

```{r}
mo$table[, c("doi", "replication_doi", "replication_url")] |>
  knitr::kable()
```

## Option: show replication outcomes

By default the module reports *that* a replication exists. Set `show_outcomes = TRUE` to also include the outcome of each replication (for example, whether it was successful), which adds the `replication_outcome` and `replication_type` information to the report:

```{r}
mo <- module_run(paper, "ref_replication", show_outcomes = TRUE)
mo$table[, c("doi", "replication_doi", "replication_outcome", "replication_type")] |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "ref_replication", show_outcomes = TRUE)
mo$summary_table
```

## Interpreting the result

A match means a replication of a cited study exists in the database — it is an invitation to check whether the replication's result changes how you should cite the original. As always, the module surfaces information for human judgement.

## Options

| Argument | Default | Effect |
|---|---|---|
| `paper` | — | the paper or paperlist to check |
| `show_outcomes` | `FALSE` | include the outcome and type of each replication |

## Updating the database

The module matches references against a data frame called `FLoRA` that ships with Metacheck. The FLoRA replication database grows as new replications are recorded, so the bundled copy gradually goes out of date. Two functions help you manage it:

```{r}
#| eval: false
FLoRA_date()     # the date the bundled database was last updated
FLoRA_update()   # download the newest version of the FLoRA database
```

`FLoRA_update()` downloads the latest FLoRA data (>5 MB), summarises it into a compact form, deletes the large original, and caches the result so subsequent checks use the fresh data. Run it occasionally to pick up replications added since your installed version of Metacheck.
