---
title: "Ethics Approval Check"
vignette: >
  %\VignetteIndexEntry{Ethics Approval Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `ethics_check` module searches a paper for a statement that the research received **ethics approval** ŌĆö for example from an Institutional Review Board (IRB), a Research Ethics Committee, or an animal-care committee ŌĆö or that it was explicitly exempted or waived. Reporting ethics approval is expected for studies involving human or animal participants.

The module recognises a wide range of committee names and approval phrasing across several languages and conventions, including:

- **Human research committees** ŌĆö Institutional Review Board (IRB), Independent Ethics Committee (IEC), Research/Medical Ethics Committee, and abbreviations such as REC, REB, ERB, METC, DEC.
- **Animal research committees** ŌĆö IACUC, Animal Welfare Ethical Review Body (AWERB), and language variants.
- **Approval phrasing** ŌĆö "approved by the ŌĆ” ethics committee", "ethically approved", "approved under protocol ŌĆ”".
- **Waivers and exemptions** ŌĆö "ethics waiver", "deemed exempt", "IRB exemption".
- **Declaration of Helsinki** statements.

This module is **fully offline** ŌĆö it uses only regular expressions, so all the code below runs with just the metacheck package.

## Running the module

The module only flags a *missing* statement when a paper appears to involve **live data collection** (so re-analyses of existing data are not penalised). We run it across the first ten papers of the built-in `psychsci` dataset.

```{r}
mo <- module_run(psychsci[1:10], "ethics_check")
mo$traffic_light
mo$summary_text
```

The `summary_table` has one row per paper, with an `ethics_approved` flag and the matched statement(s):

```{r}
mo$summary_table[, c("paper_id", "ethics_approved")] |>
  knitr::kable()
```

## A clean example and one with problems

Papers that report approval show the exact sentence the module matched ŌĆö useful for confirming it found a genuine statement:

```{r}
approved <- mo$summary_table[mo$summary_table$ethics_approved %in% TRUE, ]
approved$ethics_statements[[1]]
```

Papers that appear to collect data but contain **no** approval statement are the ones to check. These are what turn the traffic light red:

```{r}
mo$summary_table[mo$summary_table$needs_ethics %in% TRUE &
                 mo$summary_table$ethics_approved %in% FALSE, "paper_id"]
```

A missing statement is not proof that approval was never obtained ŌĆö many authors simply omit it ŌĆö but it flags papers where a reviewer should confirm the ethics reporting.

## Options

`ethics_check` takes only the `paper` argument.

## Notes

Because the module matches committee names and approval phrasing with regular expressions, it can miss unusually worded statements, and may occasionally match a sentence that mentions ethics for another reason. As with every module, treat the output as a prompt for human judgement rather than a verdict.
