---
title: "Codebook Check"
vignette: >
  %\VignetteIndexEntry{Codebook Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

Shared data is only reusable if others can tell what each column *means*. The `codebook_check` module measures **documentation coverage**: for every data column extracted by [`data_check`](mod-data-check.qmd), it asks whether that column is described in a codebook (a data dictionary) shared alongside the data.

It reports, for each column, one of these statuses:

- **labelled** — the column is matched to an entry in a codebook file.
- **unlabelled** — no documentation was found for the column.
- **conflicted** — the codebook contains more than one, disagreeing, description.
- **llm** — the label was inferred by an LLM (only when one is enabled).

It also reports **unused** codebook entries: variables documented in the codebook that do not correspond to any column in the data.

::: callout-note
This chapter runs offline against a small folder containing a data file and a codebook.
:::

## A worked example

We create a study with a `codebook.csv` documenting two of its three columns:

```{r}
repo <- file.path(tempdir(), "codebook_check_demo")
dir.create(file.path(repo, "data"), recursive = TRUE, showWarnings = FALSE)

utils::write.csv(
  data.frame(id = 1:6,
             rt = c(510, 490, 505, 512, 498, 530),
             cond = c(0, 1, 0, 1, 0, 1)),
  file.path(repo, "data", "study.csv"), row.names = FALSE)

# a codebook documenting rt and cond (but not id)
writeLines(c("variable,description",
             "rt,reaction time in milliseconds",
             "cond,condition: 0 = control, 1 = treatment"),
           file.path(repo, "codebook.csv"))
```

```{r}
mo <- module_run(test_paper(), "codebook_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## The coverage table

The `table` gives one row per data column with its documentation status and, where matched, the label found:

```{r}
mo$table[, c("column_name", "label_status", "label", "codebook_variable")] |>
  knitr::kable()
```

Here `rt` and `cond` are **labelled** (matched to the codebook), while `id` is **unlabelled** — there was no codebook entry for it.

Matching is not purely literal: names are normalised (case, separators, and simple stemming) so that `reaction_time` in the data matches `Reaction Time` in the codebook. This tolerance is important because codebooks and data files rarely use byte-identical names.

## The summary counts

```{r}
mo$summary_table |>
  knitr::kable()
```

| Column | Meaning |
|---|---|
| `column_n` | total data columns |
| `matched_n` | columns matched to a codebook entry |
| `unmatched_n` | columns with no documentation |
| `clean_n` / `conflicted_n` | matched columns with a single vs. a contradictory definition |
| `codebook_var_n` | entries found in the codebook(s) |
| `unused_var_n` | codebook entries not used by any column |

## The traffic light

| Light | Meaning |
|---|---|
| **green** | all (or nearly all) columns are documented |
| **yellow** | some columns are undocumented |
| **red** | no codebook was found, or documentation coverage is very low |

## Using an LLM

Many repositories document their variables in prose — a README paragraph, a methods section, or a non-tabular codebook — rather than a tidy `variable,description` table. With `llm_use(TRUE)`, `codebook_check` can read those unstructured sources and propose labels for otherwise unlabelled columns, marking them with the **llm** status so you can tell inferred labels from author-supplied ones. For large repositories the LLM calls are batched to stay within request limits.

```{r}
#| eval: false
llm_use(TRUE)
llm_model("groq/openai/gpt-oss-20b")
module_run(test_paper(), "codebook_check",
           local_path = repo, local_only = TRUE)
```

See the [LLMs](llms.qmd) chapter for model configuration.

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `green` / `yellow` / `red` / `na` |
| `$summary_text` | plain-text coverage summary |
| `$summary_table` | the counts described above |
| `$table` | one row per column: `column_name`, `label`, `label_status`, `codebook_variable`, ... |
| `$report` | formatted report of documented vs. undocumented columns |

## Related

- [Data Check](mod-data-check.qmd) supplies the columns being documented.
- The `convert_codebook()` function turns this output into a fully **labelled** data frame and a runnable codebook document — see the [Creating a Report](creating-a-report.qmd) workflow.
- [Psych-DS Check](mod-psychds-check.qmd) checks the repository's *structure* against a standard, complementing this check of its *documentation*.
