---
title: "Repository Check"
vignette: >
  %\VignetteIndexEntry{Repository Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `repo_check` module finds links to **research repositories** in a paper (OSF, GitHub, ResearchBox, Zenodo), retrieves the list of files in each, and reports what is shared — for example whether there are data and code files, and whether the repository includes a README.

::: callout-note
This module makes **live network calls** to the linked repositories. You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` links to several repositories.

```{r}
paper <- demopaper()
mo <- module_run(paper, "repo_check")
mo$traffic_light
cat(mo$summary_text)
```

The `table` lists every file found across all linked repositories:

```{r}
mo$table[, c("repo_name", "file_name", "file_type", "file_size")] |>
  head(10) |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "repo_check")
mo$summary_table
```

## A clean example and one with problems

A repository that includes a README and shared data/code is the "good" case; repositories with no README, or no shared files, are surfaced for attention. The summary text reports both:

```{r}
cat(mo$summary_text)
```

## Options

`repo_check` accepts arguments for working with **local files** instead of (or in addition to) online repositories:

```{r}
#| eval: false
# also include files from a local folder
module_run(paper, "repo_check", local_path = "path/to/downloaded/files")

# only check a local folder, skip all online lookups
module_run(paper, "repo_check", local_path = "path/to/files", local_only = TRUE)
```

`local_only = TRUE` is useful when you cannot or do not want to contact external services — for example when checking a reviewer submission you downloaded as a zip. See the [Local Files](local-files.qmd) chapter for details.

## Related

The companion [code_check](mod-code-check.qmd) module analyses the actual code files that `repo_check` discovers.
