---
title: "Exact p-Values"
vignette: >
  %\VignetteIndexEntry{Exact p-Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `stat_p_exact` module flags **p-values reported with insufficient precision** — for example `p < .05` or `p = n.s.` instead of an exact value like `p = .031` — and p-values reported as **exactly zero** (`p = .000`), which is mathematically impossible and should be written `p < .001`.

Reporting exact p-values is an APA Journal Article Reporting Standard (JARS). It also lets readers include results in p-curve or z-curve meta-analyses and check internal consistency with tools like Statcheck.

This module is **fully offline**.

## Running the module

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "stat_p_exact")
```

The `table` lists every detected p-value, with `imprecise` and `zero` flags:

```{r}
mo <- module_run(paper, "stat_p_exact")
mo$traffic_light
mo$summary_text
mo$table[, c("text", "p_comp", "p_value", "imprecise", "zero")] |>
  knitr::kable()
```

To see only the problematic ones:

```{r}
mo$table[mo$table$imprecise | mo$table$zero, c("text", "expanded")] |>
  knitr::kable()
```

## Running on many papers

```{r}
mo <- module_run(psychsci[1:20], "stat_p_exact")
head(mo$summary_table) |>
  knitr::kable()
```

## A clean example and one with problems

A known paper that reports imprecise p-values:

```{r}
#| results: asis
module_run(psychsci$`0956797614560771`, "stat_p_exact")
```

A paper that reports all p-values exactly returns green:

```{r}
#| results: asis
module_run(psychsci$`0956797616665351`, "stat_p_exact")
```

## Options

`stat_p_exact` takes only the `paper` argument.

## Validation

> In a sample of 225 papers containing 405 instances of non-exact p-values, the module correctly detected 269 cases (true positives) and incorrectly identified 78 (false positives). It missed 136 instances (false negatives) and correctly identified 4557 cases of precisely reported p-values (true negatives). 78% of positive detections were correct (positive predictive value).
