---
title: "Excel Check"
vignette: >
  %\VignetteIndexEntry{Excel Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

Excel files are convenient to make but easy to make in ways that lose information when the data are read by a program or exported to CSV. The `excel_check` module inspects every `.xlsx` file shared with a paper and flags four kinds of **non-machine-readable formatting**:

- **Colour coding** — cells whose *fill colour* is used to encode information (for example, marking excluded participants in red). Colour is invisible to any analysis script and is lost on CSV export.
- **Merged cells** — merged ranges break the rectangular grid that data frames require, and typically indicate multi-level headers that a program cannot parse.
- **Empty rows** — fully blank rows inside the data range.
- **Empty or unnamed columns** — columns with a blank header, or a header with no data beneath it.

None of these are errors in Excel itself; they are problems only when the file is meant to be *shared data*. `excel_check` surfaces them so the author can flatten the workbook into a clean, rectangular, machine-readable table.

::: callout-note
This chapter builds example `.xlsx` files with the `openxlsx` package. If you do not have it installed, the code chunks will not run, but the explanations still apply.
:::

## A worked example

We make a folder with one messy workbook (colour, a merge, and an empty column) and one clean one:

```{r}
#| eval: !expr requireNamespace("openxlsx", quietly = TRUE)
repo <- file.path(tempdir(), "excel_check_demo")
dir.create(file.path(repo, "data"), recursive = TRUE, showWarnings = FALSE)

# Messy workbook
wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "Data")
df <- data.frame(id = 1:4, group = c("a", "b", "a", "b"),
                 notes = rep(NA, 4), score = c(10, 20, 30, 40))
openxlsx::writeData(wb, "Data", df)
openxlsx::addStyle(wb, "Data", openxlsx::createStyle(fgFill = "#FFCC00"),
                   rows = 2, cols = 4)                 # a colour-coded cell
openxlsx::mergeCells(wb, "Data", cols = 1:2, rows = 7) # a merged range
openxlsx::saveWorkbook(wb, file.path(repo, "data", "messy.xlsx"),
                       overwrite = TRUE)

# Clean workbook
openxlsx::write.xlsx(data.frame(id = 1:3, score = c(1.1, 2.2, 3.3)),
                     file.path(repo, "data", "clean.xlsx"))
```

```{r}
#| eval: !expr requireNamespace("openxlsx", quietly = TRUE)
mo <- module_run(test_paper(), "excel_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## The findings table

Each issue is reported as one row naming the file, the worksheet, the issue type, and a human-readable detail:

```{r}
#| eval: !expr requireNamespace("openxlsx", quietly = TRUE)
mo$table |>
  knitr::kable()
```

Notice that only `messy.xlsx` is flagged — the clean workbook produces no rows. The `summary_table` counts each issue type across all files (`color_n`, `merge_n`, `empty_row_n`, `empty_col_n`) alongside how many files were inspected and how many had at least one problem.

## The traffic light

| Light | Meaning |
|---|---|
| **green** | Excel files were found and none had formatting problems |
| **yellow** | at least one file has colour coding, merges, empty rows, or empty columns |
| **na** | no Excel files were found |

## A note on `.xls` and speed

- **Legacy `.xls`** files cannot be inspected for colour and merges the same way; they are reported as *"un-inspectable"* with a suggestion to convert to `.xlsx`.
- `excel_check` reads the workbook's XML directly to find colour and merge information that `readxl` discards. This is done efficiently even for very wide survey exports (thousands of columns), so it stays fast on large files.

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `green` / `yellow` / `na` |
| `$summary_text` | plain-text summary of the issues found |
| `$summary_table` | `file_n`, `flagged_file_n`, `color_n`, `merge_n`, `empty_row_n`, `empty_col_n` |
| `$table` | one row per issue: `File`, `Sheet`, `Issue`, `Detail` |
| `$report` | formatted report listing the issues per file |

## Related

- [Data Check](mod-data-check.qmd) reads the *contents* of Excel files; `excel_check` inspects their *formatting*, which is lost once a file is read into a data frame.
- [Data Validate](mod-data-validate.qmd) checks the values inside the columns.
