---
title: "Power Analysis Check"
vignette: >
  %\VignetteIndexEntry{Power Analysis Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `power` module uses regular expressions to find sentences that describe a statistical **power analysis**, and classifies each as *a priori*, *sensitivity*, or *post-hoc*. When [LLM support](llms.html) is turned on, it additionally reads each power sentence and extracts structured details — the statistical test, planned sample size, alpha level, desired power, and effect size — and checks whether the analysis is fully reported.

A priori power analyses are the useful kind; post-hoc ("observed") power is rarely informative.

## Running the module

Every module is run the same way, with `module_run(paper, "module_name")`. We use the built-in `demopaper()`, which contains a few power statements.

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "power")
```

Printing the result renders the module's report. To work with the underlying data, use the returned list. The `table` has one row per detected power statement, with a `power_type` classification and a `complete` flag:

```{r}
mo <- module_run(paper, "power")
mo$traffic_light
mo$summary_text
mo$table[, c("text", "power_type", "complete")] |>
  knitr::kable()
```

## Running on many papers

Pass a `paperlist` such as the built-in `psychsci` (250 open-access *Psychological Science* articles) to run across a corpus. Here we run on the first 20 papers and summarise how many power statements were found per paper.

```{r}
mo <- module_run(psychsci[1:20], "power")
head(mo$summary_table) |>
  knitr::kable()
```

## A clean example and one with problems

A paper that reports an a priori power analysis with all the needed details is the "good" case; a paper that mentions power only vaguely, or reports post-hoc power, is the kind of thing the module surfaces for a closer look.

```{r}
#| results: asis
# demopaper contains both a priori and less-complete power statements
module_run(demopaper(), "power")
```

The `power_type` column tells you which statements are a priori (worth keeping) versus post-hoc (usually a red flag), and `complete` flags a priori analyses that are missing required information.

## Options: using an LLM

The regular-expression behaviour above runs with no setup. To also **extract structured details** — the statistical test, sample size, alpha, power, and effect size — turn on [LLM support](llms.html). The module reads the same switch (`llm_use()`) and model (`llm_model()`) described in the LLM chapter; there is no special module argument. When an LLM is enabled, the result table gains columns such as `statistical_test`, `sample_size`, `power`, `effect_size`, and `effect_size_metric`.

Below we run exactly the same call three ways — with no LLM, with a local Ollama model, and with a cloud model on Groq — on `demopaper()`, and compare what each returns. (The LLM outputs shown here were captured from real runs; running the LLM at build time would require Ollama or an API key, so the results are loaded from saved objects.)

### No LLM (default)

```{r}
#| eval: false
llm_use(FALSE)
module_run(demopaper(), "power")
```

This is the regex-only behaviour shown earlier: it detects and classifies power statements, but the extraction columns are empty.

### Local LLM with Ollama

This run used `qwen3.5:9b` (a 9.7-billion-parameter model) running locally with Ollama. Nothing leaves your machine, but the model does the work, so it is slower (this call took about **105 seconds**).

```{r}
#| eval: false
llm_use(TRUE)
llm_model("ollama/qwen3.5:9b")
module_run(demopaper(), "power")
```

```{r}
#| echo: false
ollama_power <- readRDS("_power_ollama.Rds")
ollama_power$summary_text
ollama_power$table[, c("power_type", "statistical_test", "sample_size",
                       "power", "effect_size", "effect_size_metric", "complete")] |>
  knitr::kable()
```

The local model found the one genuine power analysis in the demo paper — the sensitivity analysis — and correctly extracted Cohen's *d* = 0.5 and 80% power. It did not flag anything else.

### Cloud LLM with Groq

This run used Groq's default model (`llama-3.1-8b-instant`). It is much faster (about **11 seconds**) but sends the text to an external service.

```{r}
#| eval: false
llm_use(TRUE)
llm_model("groq")
module_run(demopaper(), "power")
```

```{r}
#| echo: false
groq_power <- readRDS("_power_groq.Rds")
groq_power$summary_text
groq_power$table[, c("power_type", "statistical_test", "sample_size",
                     "power", "effect_size", "effect_size_metric", "software", "complete")] |>
  knitr::kable()
```

### Discussing the difference

The two models behaved noticeably differently on the same input:

- **Recall vs. precision.** The local `qwen3.5:9b` returned a single, clean row — the real sensitivity analysis. Groq's smaller, faster `llama-3.1-8b-instant` returned three rows: it caught the same sensitivity analysis, but also flagged the paper's abstract (a **false positive**, with every field `NA`) and a line of `pwr.t.test()` example code (extracting `software = R` and `statistical_test = t-test`).
- **Extraction detail.** Both correctly pulled Cohen's *d* = 0.5 and power = 0.8 for the sensitivity analysis. Neither marked any analysis as `complete`, because the demo paper does not report every required field.
- **Speed vs. privacy.** Groq returned in ~11 s versus ~105 s locally, but the local run kept the manuscript on the machine. For unpublished work, the slower local model is usually the right trade-off.

There is no single "correct" answer here — the output depends on the model. A larger or more capable model generally gives cleaner extraction with fewer false positives, which is exactly why the model you choose matters. This is also why module output is meant to **prompt human judgement** rather than serve as a verdict: a person reading these tables quickly sees that the abstract and the code line are not real power analyses. See the [Local AI with Ollama](ollama.html) chapter for more on choosing and running local models.

### Reproducibility

The `power` module accepts a `seed` argument (default `8675309`) passed to the LLM, so repeated runs with the same model are more reproducible:

```{r}
#| eval: false
module_run(demopaper(), "power", seed = 1)
```

## Validation

> In a sample of 128 papers with 246 instances of power statements, 203 were correctly detected (true positives), 22 were missed (false negatives) and 21 were incorrectly detected (false positives). Among all instances flagged as power statements, 90.6% were correct (positive predictive value).

As with all modules, treat the output as a prompt for human judgement, not a verdict.
