---
title: "Psych-DS Check"
vignette: >
  %\VignetteIndexEntry{Psych-DS Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

[Psych-DS](https://psych-ds.github.io/) is a community standard for organising a psychology dataset so that both humans and machines can find their way around it: data live in a `data/` folder, analysis code in `analysis/`, materials in `materials/`, and a machine-readable `dataset_description.json` at the root describes the dataset and its variables.

The `psychds_check` module compares a repository against this standard and reports the **compliance gap**: for every file, where it currently sits, where the Psych-DS layout says it should go, and what (if anything) is missing at the top level.

It answers three questions:

- Are the **required** pieces present (a `dataset_description.json`, a `data/` directory)?
- Are the **recommended** pieces present (a README, a `CHANGES` file)?
- Which files are **misplaced** relative to the standard, and where should they move?

::: callout-note
This chapter runs offline against a small repository that does *not* follow the standard, so we can see the gap the module reports.
:::

## A worked example

```{r}
repo <- file.path(tempdir(), "psychds_check_demo")
dir.create(repo, recursive = TRUE, showWarnings = FALSE)

# a flat repo: data and code sitting at the top level, no metadata
utils::write.csv(data.frame(id = 1:5, score = rnorm(5)),
                 file.path(repo, "study_data.csv"), row.names = FALSE)
writeLines("# analysis\n", file.path(repo, "analysis.R"))
writeLines("A short readme.", file.path(repo, "README.md"))
```

```{r}
mo <- module_run(test_paper(), "psychds_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## The compliance table

The `table` lists each file with its current path and the Psych-DS target path, plus a `status` saying what needs to happen:

```{r}
mo$table[, c("file_name", "data_type", "current_path", "target_path", "status")] |>
  knitr::kable()
```

A `status` of `move` means the file is in the wrong place for a Psych-DS dataset — for example `study_data.csv` should move into `data/`, and `analysis.R` into `analysis/`. This gives the author a concrete to-do list rather than an abstract "does not comply".

## The target tree

The most useful part of the report is the **target tree**: a drawing of the Psych-DS layout the repository *should* have, with the current files placed where they belong and the missing pieces marked. It turns the compliance table into a single picture you can act on.

The tree uses colour and annotations to show three states:

- **files in <span style="color:#c0392b">red</span>** with a *← missing* marker are required by Psych-DS but absent (for example `dataset_description.json` and `CHANGES`);
- **files annotated <span style="color:#b9770e">(move from ...)</span>** exist in the repository but need relocating into the standard folder — the annotation tells you exactly where they are now;
- **plain files** are already in the right place.

The tree is emitted as HTML in the module's report. Here it is for our example repository:

```{r}
#| results: asis
tree_html <- grep("<pre", mo$report, value = TRUE)
cat(tree_html)
```

Reading this top to bottom is a complete migration plan: create the two missing metadata files, move the data file into `data/` (renamed to the Psych-DS `source-*` convention) and the script into `analysis/`, and the repository will comply. The `convert_psychds()` function performs exactly these steps automatically.

## The summary counts

```{r}
mo$summary_table |>
  knitr::kable()
```

| Column | Meaning |
|---|---|
| `required_met` / `required_missing` | required Psych-DS elements present / absent |
| `recommended_met` / `recommended_missing` | recommended elements present / absent |
| `misplaced_n` | files that are not in their Psych-DS location |

## The traffic light

| Light | Meaning |
|---|---|
| **green** | the repository already follows the Psych-DS layout |
| **yellow** | some elements are missing or files are misplaced |
| **red** | required elements (like `data/` or the description file) are absent |
| **na** | there were no files to assess |

## From gap report to a compliant dataset

`psychds_check` tells you *what* is missing; the companion function `convert_psychds()` acts on it. Given the same repository, it writes out a Psych-DS-compliant copy — moving files into `data/`, `analysis/`, and so on, and generating the `dataset_description.json` metadata from the columns `data_check` extracted:

```{r}
#| eval: false
res <- report(paper, c("data_check", "psychds_check"))
convert_psychds(res, overwrite = TRUE)
```

The generated dataset is validated with `psychds_validate()`, Metacheck's built-in checker for the standard, so the converted output is guaranteed to pass.

## Building a whole data archive

The same machinery scales up from one repository to a **data release** — a browsable, archivable copy of the data behind an entire corpus of papers, each converted to Psych-DS. This is a good way to turn scattered OSF/GitHub/Zenodo repositories into a single, structured, machine-readable collection.

The recipe is: run the data pipeline over each paper with `data_check` set to fetch **all** files, then convert each to Psych-DS. Two `data_check` options (covered in the [Data Check](mod-data-check.qmd) chapter) make this practical for real-world repositories:

- `download = "all"` fetches every file, not just the readable data.
- `skip_types = "asset"` leaves out stimuli/media (which a release links to rather than hosts), and `peek_zips = TRUE` looks inside zips and only downloads those containing data.

```{r}
#| eval: false
for (paper in papers) {
  chain <- report(paper,
                  c("data_check", "codebook_check", "psychds_check"),
                  args = list(data_check = list(
                    download = "all", skip_types = "asset", peek_zips = TRUE)))
  convert_psychds(chain,  output_dir = file.path("archive", paper_id(paper), "psychds"))
  convert_codebook(chain, output_dir = file.path("archive", paper_id(paper), "codebook"))
}
```

### Zips are unpacked into the archive

When `peek_zips = TRUE`, a downloaded zip that contains data is unpacked and its inner data files are placed into the archive individually — so a `data.zip` becomes real, browsable `data/` files rather than an opaque bundle. Two helper functions expose this directly:

- **`zip_peek(url)`** reads a remote zip's file listing (names and sizes) via an HTTP range request, *without* downloading it.
- **`zip_decision(url)`** peeks and classifies the contents, returning whether the zip is worth downloading (contains data) or should just be linked (only stimuli).

### An archive-level catalogue

A release of many datasets needs a top-level description tying them together. **`data_catalog()`** scans a built archive directory and writes a `catalog.json` — a schema.org [`DataCatalog`](https://schema.org/DataCatalog) listing every dataset with its title, authors, DOI, and file count — plus a `catalog.csv` for quick inspection. This is the object that makes the release a single discoverable archive (and is harvested by Google Dataset Search), rather than a set of unconnected folders.

```{r}
#| eval: false
data_catalog("archive", papers = papers, name = "My data release")
```

::: callout-tip
For a large corpus, first run `data_check` with `download = "none"` and `manifest = "..."` to write a per-paper file manifest (file names, sizes, types) *without* downloading anything. Inspecting those manifests lets you choose the size caps knowing exactly which files each value would include or exclude, before committing to a large download.
:::

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `green` / `yellow` / `red` / `na` |
| `$summary_text` | plain-text compliance summary |
| `$summary_table` | the required/recommended/misplaced counts above |
| `$table` | one row per file: `file_name`, `data_type`, `current_path`, `target_path`, `status` |
| `$report` | formatted report, including the target directory tree with missing pieces marked |

## Related

- [Data Check](mod-data-check.qmd) classifies the files that `psychds_check` re-organises.
- [Codebook Check](mod-codebook-check.qmd) checks variable *documentation*, which becomes the `variableMeasured` metadata in a Psych-DS dataset.
- See the [OSF Prereg Schemas](osf-schemas.qmd) appendix for related standards Metacheck understands.
