---
title: "Repository Check"
vignette: >
  %\VignetteIndexEntry{Repository Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `repo_check` module finds links to **research repositories** in a paper (OSF, GitHub, ResearchBox, Zenodo), retrieves the list of files in each, and reports what is shared — for example whether there are data and code files, and whether the repository includes a README.

::: callout-note
This module makes **live network calls** to the linked repositories. You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` links to several repositories.

```{r}
paper <- demopaper()
mo <- module_run(paper, "repo_check")
mo$traffic_light
cat(mo$summary_text)
```

The `table` lists every file found across all linked repositories:

```{r}
mo$table[, c("repo_name", "file_name", "file_type", "file_size")] |>
  head(10) |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "repo_check")
mo$summary_table
```

## What it checks for

For every repository it finds, `repo_check` builds a unified file list and then assesses three things:

- **A README.** Each repository should document its contents. The module looks for a file whose name contains "readme" (or "read me"). A repository without one is flagged, because readers have no map of what the shared files are.
- **Archive (zip) files.** Files bundled into a `.zip` (or similar archive) are flagged, because their contents cannot be inspected, indexed, or reused without downloading and unpacking them. The module reports the archive names and suggests uploading the files individually.
- **What is shared.** It counts the data files, code files, README files, and archives in each repository (the `files_data`, `files_code`, `files_readme`, and `files_zip` columns of the summary table), so you can see at a glance whether data and code were actually shared.

The traffic light follows from these checks:

- **green** — repositories were found, every one has a README, and there are no archive files.
- **yellow** — at least one repository is missing a README, or one or more archive files were found.
- **na** — no repository links were found in the paper at all.

## A clean example and one with problems

The demo paper links to three repositories, and running `repo_check` on it produces a **yellow** light — there are problems to address. The summary text reports exactly what:

```{r}
mo$traffic_light
cat(mo$summary_text)
```

Here the issues are a **missing README** and an **archive file** whose contents could not be examined. The per-paper summary table counts each file type — note `files_readme` is 0 (no README was shared) and `files_zip` is 1 (one archive):

```{r}
mo$summary_table |>
  knitr::kable()
```

We can see the specific archive file that was flagged in the full file table:

```{r}
mo$table[mo$table$file_type %in% "archive", c("repo_name", "file_name")] |>
  knitr::kable()
```

A **clean** result, by contrast, would include a README in every linked repository (`files_readme` ≥ 1 for each) and share files individually rather than as a zip (`files_zip` of 0) — producing a green light with no items to address.

## Options

`repo_check` accepts arguments for working with **local files** instead of (or in addition to) online repositories:

```{r}
#| eval: false
# also include files from a local folder
module_run(paper, "repo_check", local_path = "path/to/downloaded/files")

# only check a local folder, skip all online lookups
module_run(paper, "repo_check", local_path = "path/to/files", local_only = TRUE)
```

`local_only = TRUE` is useful when you cannot or do not want to contact external services — for example when checking a reviewer submission you downloaded as a zip. See the [Local Files](local-files.qmd) chapter for details.

## Related

The companion [code_check](mod-code-check.qmd) module analyses the actual code files that `repo_check` discovers.
