---
title: "Preregistration Check"
vignette: >
  %\VignetteIndexEntry{Preregistration Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `prereg_check` module finds links to **preregistrations** on the OSF and AsPredicted in the manuscript, retrieves the preregistration text, and organises its contents into a single standardised template. This makes it much easier for a reviewer to compare what was preregistered against what the manuscript reports.

::: callout-note
This module makes **live network calls** to the OSF and AsPredicted to retrieve preregistration content. You need an internet connection to run the code below. (When this book is built, these calls run against the live services.)
:::

## Running the module

`demopaper()` links to two preregistrations, so it is a good example.

```{r}
paper <- demopaper()
mo <- module_run(paper, "prereg_check")
mo$traffic_light
mo$summary_text
```

The `table` has one row per preregistration, with the template fields organised into columns. Here are a few of them:

```{r}
mo$table[, c("template_name", "title", "id", "sample_size")] |>
  knitr::kable()
```

The full set of fields the module standardises includes the study design, hypotheses, sample-size rationale, stopping rule, variables, statistical tests, inference criteria, and exclusion criteria:

```{r}
names(mo$table)
```

## Reading a single preregistration's contents

Because each preregistration becomes a row, you can inspect one in detail by transposing it:

```{r}
prereg <- mo$table[1, c("template_name", "sample_size", "statistical_tests", "data_exclusion_criteria")]
t(prereg) |>
  knitr::kable()
```

## Options and limitations

`prereg_check` takes only the `paper` argument. It **cannot** extract information from:

- unstructured preregistration templates (where everything is in one free-text field), or
- preregistrations stored as uploaded text documents on the OSF rather than in the structured form.

For the details of which OSF registration templates exist and how they are identified, see the [OSF Registration Schemas](osf-schemas.qmd) chapter.

## Notes

If you want Metacheck to support additional preregistration platforms, reach out to the development team.
