---
title: "Reference Summary"
vignette: >
  %\VignetteIndexEntry{Reference Summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `ref_summary` module does not run a check of its own. Instead it **collects the results of the other reference-section modules** — [accuracy](mod-ref-accuracy.qmd), [PubPeer](mod-ref-pubpeer.qmd), [replication](mod-ref-replication.qmd), and [retraction](mod-ref-retraction.qmd) — into a single table with one row per reference, so you can see everything known about each cited work in one place.

Because it summarises the output of those modules, `ref_summary` must run **after** them, in the same report. On its own it has nothing to summarise.

::: callout-note
The modules it summarises make **live network calls**, so an internet connection is needed.
:::

## Running the module

The summary needs the other reference modules to have run first. The natural way to arrange that is to run them together with `report_module_run()`, which keeps each module's output and passes the chain along (see [Creating a Metacheck Report](creating-a-report.qmd)). List `ref_summary` last:

```{r}
paper <- demopaper()
module_output <- report_module_run(
  paper,
  modules = c("ref_accuracy", "ref_pubpeer",
              "ref_replication", "ref_retraction",
              "ref_summary")
)

mo <- module_output[["ref_summary"]]
mo$traffic_light
mo$summary_text
```

The combined `table` has one row per reference and one column per upstream check:

```{r}
mo$table[, c("bib_id", "accuracy_mismatch", "pubpeer",
             "replication_type", "retractionwatch")] |>
  knitr::kable()
```

A reference is interesting when any of these columns is filled in: a reported accuracy mismatch, a PubPeer comment, a replication outcome, or a Retraction Watch entry. Follow up those references using the dedicated chapter for whichever column flagged it.

## Interpreting the result

`ref_summary` inherits everything from the modules it summarises, so the same cautions apply: a filled-in cell is a prompt to look, not a verdict. Its value is convenience — instead of reading four separate tables, you scan one. The summary itself is informational (`info`); the underlying modules carry the traffic lights.

## Options

`ref_summary` takes only the `paper` argument, but it only produces useful output when the other `ref_*` modules have run before it in the same report.
