---
title: "Codebook Check"
vignette: >
  %\VignetteIndexEntry{Codebook Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

Shared data is only reusable if others can tell what each column *means*. The `codebook_check` module measures **documentation coverage**: for every data column extracted by [`data_check`](mod-data-check.qmd), it asks whether that column is described in a codebook (a data dictionary) shared alongside the data.

It reports, for each column, one of these statuses:

- **labelled** — the column is matched to an entry in a codebook file.
- **unlabelled** — no documentation was found for the column.
- **conflicted** — the codebook contains more than one, disagreeing, description.
- **llm** — the label was inferred by an LLM (only when one is enabled).

It also reports **unused** codebook entries: variables documented in the codebook that do not correspond to any column in the data.

::: callout-note
This chapter runs offline against a small folder containing a data file and a codebook.
:::

## A worked example

We create a study with a `codebook.csv` documenting two of its three columns:

```{r}
repo <- file.path(tempdir(), "codebook_check_demo")
dir.create(file.path(repo, "data"), recursive = TRUE, showWarnings = FALSE)

utils::write.csv(
  data.frame(id = 1:6,
             rt = c(510, 490, 505, 512, 498, 530),
             cond = c(0, 1, 0, 1, 0, 1)),
  file.path(repo, "data", "study.csv"), row.names = FALSE)

# a codebook documenting rt and cond (but not id)
writeLines(c("variable,description",
             "rt,reaction time in milliseconds",
             "cond,condition: 0 = control, 1 = treatment"),
           file.path(repo, "codebook.csv"))
```

```{r}
mo <- module_run(test_paper(), "codebook_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## The coverage table

The `table` gives one row per data column with its documentation status and, where matched, the label found:

```{r}
mo$table[, c("column_name", "label_status", "label", "codebook_variable")] |>
  knitr::kable()
```

Here `rt` and `cond` are **labelled** (matched to the codebook), while `id` is **unlabelled** — there was no codebook entry for it.

Matching is not purely literal: names are normalised (case, separators, and simple stemming) so that `reaction_time` in the data matches `Reaction Time` in the codebook. This tolerance is important because codebooks and data files rarely use byte-identical names.

## Richer variable metadata

A good codebook records more than a one-line label. Following the [DDI](https://ddialliance.org/) standard, `codebook_check` also captures four further per-variable properties whenever a source supplies them, and carries them onto the matched data columns:

| Column | DDI concept | What it holds |
|---|---|---|
| `value_labels` | CodeList / ValueDomain | the code → label mapping, e.g. `1 = "Male", 2 = "Female"` |
| `missing_values` | MissingValues | which codes denote missingness (e.g. `-99 = "Refused"`) |
| `question` | QuestionText | the question the respondent was asked |
| `universe` | Universe | the population / filter the variable applies to |

These come from **two sources for free**. From SPSS/Stata/SAS files, `codebook_check` reads the embedded **value labels** and **declared missing values** that haven attaches to each column — information most tools discard. From a text codebook, it parses a *"values"* / *"coding"* column (understanding common encodings like `1 = Male; 2 = Female` or `0: no | 1: yes`) and dedicated *question* and *universe* columns.

Let's document `sex` with a coding column and a question:

```{r}
repo2 <- file.path(tempdir(), "codebook_values_demo")
dir.create(file.path(repo2, "data"), recursive = TRUE, showWarnings = FALSE)

utils::write.csv(
  data.frame(sex = c(1, 2, 1, 2, -99), age = c(20, 30, 40, 50, 25)),
  file.path(repo2, "data", "study.csv"), row.names = FALSE)

writeLines(c("variable,description,values,question",
             "sex,Participant sex,1 = Male; 2 = Female; -99 = Refused,What is your sex?",
             "age,Age in years,,How old are you?"),
           file.path(repo2, "data", "codebook.csv"))

mo2 <- module_run(test_paper(), "codebook_check",
                  local_path = repo2, local_only = TRUE)
mo2$table[, c("column_name", "value_labels", "missing_values", "question")] |>
  knitr::kable()
```

The code list is stored as a compact JSON string (`{"1":"Male","2":"Female","-99":"Refused"}`); a code whose label reads as missingness (`Refused`, `N/A`, `Prefer not to answer`, …) is additionally recorded in `missing_values`, so a sentinel like `-99` is flagged as a missing code rather than a real value.

All four properties are exported into the Psych-DS metadata by `convert_psychds()`: the code list becomes a schema.org **code list** (one `PropertyValue` per code), and the missing scheme, question, and universe become namespaced `metacheck:` fields — so they travel with the converted dataset.

## The summary counts

```{r}
mo$summary_table |>
  knitr::kable()
```

| Column | Meaning |
|---|---|
| `column_n` | total data columns |
| `matched_n` | columns matched to a codebook entry |
| `unmatched_n` | columns with no documentation |
| `clean_n` / `conflicted_n` | matched columns with a single vs. a contradictory definition |
| `codebook_var_n` | entries found in the codebook(s) |
| `unused_var_n` | codebook entries not used by any column |

## The traffic light

| Light | Meaning |
|---|---|
| **green** | all (or nearly all) columns are documented |
| **yellow** | some columns are undocumented |
| **red** | no codebook was found, or documentation coverage is very low |

## Using an LLM

Many repositories document their variables in prose — a README paragraph, a methods section, or a non-tabular codebook — rather than a tidy `variable,description` table. With `llm_use(TRUE)`, `codebook_check` can read those unstructured sources and propose labels for otherwise unlabelled columns, marking them with the **llm** status so you can tell inferred labels from author-supplied ones. For large repositories the LLM calls are batched to stay within request limits.

```{r}
#| eval: false
llm_use(TRUE)
llm_model("groq/openai/gpt-oss-20b")
module_run(test_paper(), "codebook_check",
           local_path = repo, local_only = TRUE)
```

See the [LLMs](llms.qmd) chapter for model configuration.

## Identifying psychometric scales

With an LLM enabled, `codebook_check` also tries to recognise the **psychometric scales** in the data. It looks for blocks of Likert-type items — runs of adjacent integer columns (3–11 response options) that share a variable-name prefix, five items or more — and asks the model which published instrument each block is (PANAS, the Rosenberg Self-Esteem Scale, and so on), using the item wording from the codebook where available and the variable-name prefix otherwise.

Each identified scale is recorded in two new columns of the `$table`:

- **`scale`** — the instrument name (e.g. `"PANAS"`), for every item column of the block;
- **`scale_confidence`** — `high`, `medium`, or `low` (low-confidence and unrecognised blocks are left blank rather than guessed).

The report gains a **Scales** section listing what was found. When a block *looks* like a scale but could not be named, the module says so and gives concrete advice — because a scale that a language model cannot identify is usually one a human reader will struggle with too:

- name variables consistently after the instrument (`panas_1 … panas_10`) rather than `Q1`, `V3`, or `item5`;
- document the item wording in a codebook or embedded value labels — the item text is what identifies a scale;
- state the scale name and reference, the response options, and any reverse-coded items.

Without an LLM, the columns are still present but empty, and the report notes that identification was skipped.

### How scales are recorded (and why)

No metadata standard metacheck writes to has a first-class "scale" concept. [Psych-DS](https://psych-ds.github.io/) describes each variable in isolation, and schema.org — its foundation — has no property for grouping variables into a construct. The closest existing concept is DDI Codebook's variable group with `type="analysis"` ("variables combined into the same index"), but that is a separate XML format psychologists rarely use.

Rather than invent a proprietary field or emit a second file, metacheck extends the Psych-DS metadata **in place**:

- the scale **name** goes in schema.org's own [`measurementTechnique`](https://schema.org/measurementTechnique) property — its intended use is exactly to name the instrument a variable was measured with;
- the **grouping** (which variables belong to the scale, and the confidence) goes in a namespaced `metacheck:scale` / `metacheck:scales` extension, since schema.org has no native grouping and Psych-DS tolerates namespaced JSON-LD.

`convert_psychds()` and `convert_codebook()` both emit these, so the scale information travels with the converted dataset. The [Data Validate](mod-data-validate.qmd) module reuses the identified names, labelling its careless-responding findings with the real scale (e.g. "PANAS") instead of a bare column prefix.

```{r}
#| eval: false
llm_use(TRUE)
llm_model("groq/openai/gpt-oss-20b")
cc <- module_run(test_paper(), "codebook_check",
                 local_path = repo, local_only = TRUE)

# the identified scales, one row per (column, scale)
unique(cc$table[!is.na(cc$table$scale), c("column_name", "scale", "scale_confidence")])
```

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `green` / `yellow` / `red` / `na` |
| `$summary_text` | plain-text coverage summary |
| `$summary_table` | the counts described above |
| `$table` | one row per column: `column_name`, `label`, `label_status`, `codebook_variable`, `value_labels`, `missing_values`, `question`, `universe`, `scale`, `scale_confidence`, ... |
| `$report` | formatted report of documented vs. undocumented columns, plus identified scales and improvement advice |

## Related

- [Data Check](mod-data-check.qmd) supplies the columns being documented.
- The `convert_codebook()` function turns this output into a fully **labelled** data frame and a runnable codebook document — see the [Creating a Report](creating-a-report.qmd) workflow.
- [Psych-DS Check](mod-psychds-check.qmd) checks the repository's *structure* against a standard, complementing this check of its *documentation*.
