---
title: "StatCheck"
vignette: >
  %\VignetteIndexEntry{StatCheck}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `stat_check` module runs [Statcheck](https://statcheck.io/) on the paper. Statcheck recomputes the p-value from each reported test statistic and degrees of freedom, and flags cases where the reported p-value is **inconsistent** with the statistic — and, more seriously, where the inconsistency changes the significance decision (a *decision error*).

Metacheck only returns Statcheck results for **t-tests and F-tests**, the tests for which the approach has been validated [@nuijten2016]. Statcheck only detects statistics written in APA format, which is mostly used in psychology journals.

This module is **fully offline** (Statcheck runs locally).

## Running the module

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "stat_check")
```

The `table` has one row per detected test, comparing `reported_p` to `computed_p`, with `error` and `decision_error` flags:

```{r}
mo <- module_run(paper, "stat_check")
mo$traffic_light
mo$summary_text
mo$table[, c("test_type", "reported_p", "computed_p", "error", "decision_error")] |>
  knitr::kable()
```

## Running on many papers

Statcheck on a large corpus can take a while (the full `psychsci` set has more than 27,000 sentences with numbers). Here we run on a small subset:

```{r}
mo <- module_run(psychsci[1:20], "stat_check")
head(mo$summary_table) |>
  knitr::kable()
```

## A clean example and one with problems

The flagged rows are the ones to inspect — especially `decision_error == TRUE`, where the reported and recomputed p-values fall on opposite sides of .05:

```{r}
mo <- module_run(psychsci[1:20], "stat_check")
err <- mo$table[mo$table$error %in% TRUE, c("raw", "reported_p", "computed_p", "decision_error")]
head(err) |>
  knitr::kable()
```

## Options

`stat_check` takes only the `paper` argument.

::: callout-note
Statcheck treats `p = .000` as an error (you should report `p < .001`), and may misflag one-sided tests, since it assumes two-sided p-values. Always read the flagged sentence before concluding there is a real error.
:::

## Validation

> In a sample of 685 tests with 34 instances of inconsistent reporting, Statcheck correctly detected all 34 and incorrectly identified 26. So 0% of true instances were missed, and 43% of detections were false positives. See <https://osf.io/preprints/psyarxiv/tcxaj_v1/> for the full validation.

Statcheck was developed by Michèle Nuijten and Sascha Epskamp.
