---
title: "Retraction Watch (ref_retraction)"
vignette: >
  %\VignetteIndexEntry{Retraction Watch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `ref_retraction` module cross-references the paper's **cited references** against the [Retraction Watch](https://retractionwatch.com/) database. Citing a retracted article — usually unknowingly — is a common and avoidable problem. The module flags any cited reference that appears in the Retraction Watch database so you can check whether the citation is still appropriate.

::: callout-note
This module makes **live network calls** (matching references by DOI against Crossref and Retraction Watch). You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` cites a reference that is in the Retraction Watch database.

```{r}
paper <- demopaper()
mo <- module_run(paper, "ref_retraction")
mo$traffic_light
mo$summary_text
```

The `table` lists each cited reference found in the database:

```{r}
mo$table[, c("doi", "text")] |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "ref_retraction")
mo$summary_table
```

## Interpreting the result

A flagged reference is not automatically a mistake — retractions happen for many reasons, and a retracted paper is sometimes cited deliberately (for example, to discuss the retraction itself). The module's job is to make sure you are *aware*, so you can decide. Follow up each flagged DOI before acting.

## Options

`ref_retraction` takes only the `paper` argument.

## Notes

This is the same kind of check that reference managers such as [Zotero](https://www.zotero.org/blog/retracted-item-notifications/) perform, brought into the metacheck workflow so it can run alongside the other reference checks.
