---
title: "Replication Check"
vignette: >
  %\VignetteIndexEntry{Replication Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `ref_replication` module cross-references the paper's **cited references** against the [FLoRA](https://forrt.org/) replication database. If you cite an original study that has since been the subject of a replication attempt, that is highly relevant context — especially if the replication failed. The module flags cited originals that have known replications, and can report the replication's outcome.

::: callout-note
This module makes **live network calls** to the FLoRA replication database. You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` cites an original study that has a recorded replication.

```{r}
paper <- demopaper()
mo <- module_run(paper, "ref_replication")
mo$traffic_light
mo$summary_text
```

The `table` matches each cited original to its replication(s):

```{r}
mo$table[, c("doi", "replication_doi", "replication_url")] |>
  knitr::kable()
```

## Option: show replication outcomes

By default the module reports *that* a replication exists. Set `show_outcomes = TRUE` to also include the outcome of each replication (for example, whether it was successful), which adds the `replication_outcome` and `replication_type` information to the report:

```{r}
mo <- module_run(paper, "ref_replication", show_outcomes = TRUE)
mo$table[, c("doi", "replication_doi", "replication_outcome", "replication_type")] |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "ref_replication", show_outcomes = TRUE)
mo$summary_table
```

## Interpreting the result

A match means a replication of a cited study exists in the database — it is an invitation to check whether the replication's result changes how you should cite the original. As always, the module surfaces information for human judgement.

## Options

| Argument | Default | Effect |
|---|---|---|
| `paper` | — | the paper or paperlist to check |
| `show_outcomes` | `FALSE` | include the outcome and type of each replication |

## Updating the database

The module matches references against a data frame called `FLoRA` that ships with Metacheck. The FLoRA replication database grows as new replications are recorded, so the bundled copy gradually goes out of date. Two functions help you manage it:

```{r}
#| eval: false
FLoRA_date()     # the date the bundled database was last updated
FLoRA_update()   # download the newest version of the FLoRA database
```

`FLoRA_update()` downloads the latest FLoRA data (>5 MB), summarises it into a compact form, deletes the large original, and caches the result so subsequent checks use the fresh data. Run it occasionally to pick up replications added since your installed version of Metacheck.
