---
title: "Psych-DS Check"
vignette: >
  %\VignetteIndexEntry{Psych-DS Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
llm_use(FALSE)
```

::: callout-warning
This module is currently under development, and not available in the main version of Metacheck. To try it out, install the development version of Metacheck:

```r
remotes::install_github("scienceverse/metacheck@data_check")
```
:::

## What it checks

[Psych-DS](https://psych-ds.github.io/) is a community standard for organising a psychology dataset so that both humans and machines can find their way around it: data live in a `data/` folder, analysis code in `analysis/`, materials in `materials/`, and a machine-readable `dataset_description.json` at the root describes the dataset and its variables.

The `psychds_check` module compares a repository against this standard and reports the **compliance gap**: for every file, where it currently sits, where the Psych-DS layout says it should go, and what (if anything) is missing at the top level.

It answers three questions:

- Are the **required** pieces present (a `dataset_description.json`, a `data/` directory)?
- Are the **recommended** pieces present (a README, a `CHANGES` file)?
- Which files are **misplaced** relative to the standard, and where should they move?

::: callout-note
This chapter runs offline against a small repository that does *not* follow the standard, so we can see the gap the module reports.
:::

## A worked example

```{r}
repo <- file.path(tempdir(), "psychds_check_demo")
dir.create(repo, recursive = TRUE, showWarnings = FALSE)

# a flat repo: data and code sitting at the top level, no metadata
utils::write.csv(data.frame(id = 1:5, score = rnorm(5)),
                 file.path(repo, "study_data.csv"), row.names = FALSE)
writeLines("# analysis\n", file.path(repo, "analysis.R"))
writeLines("A short readme.", file.path(repo, "README.md"))
```

```{r}
mo <- module_run(test_paper(), "psychds_check",
                 local_path = repo, local_only = TRUE)

mo$traffic_light
cat(mo$summary_text)
```

## The compliance table

The `table` lists each file with its current path and the Psych-DS target path, plus a `status` saying what needs to happen:

```{r}
mo$table[, c("file_name", "data_type", "current_path", "target_path", "status")] |>
  knitr::kable()
```

A `status` of `move` means the file is in the wrong place for a Psych-DS dataset — for example `study_data.csv` should move into `data/`, and `analysis.R` into `analysis/`. This gives the author a concrete to-do list rather than an abstract "does not comply".

## The target tree

The most useful part of the report is the **target tree**: a drawing of the Psych-DS layout the repository *should* have, with the current files placed where they belong and the missing pieces marked. It turns the compliance table into a single picture you can act on.

The tree uses colour and annotations to show three states:

- **files in <span style="color:#c0392b">red</span>** with a *← missing* marker are required by Psych-DS but absent (for example `dataset_description.json` and `CHANGES`);
- **files annotated <span style="color:#b9770e">(move from ...)</span>** exist in the repository but need relocating into the standard folder — the annotation tells you exactly where they are now;
- **plain files** are already in the right place.

The tree is emitted as HTML in the module's report. Here it is for our example repository:

```{r}
#| results: asis
tree_html <- grep("<pre", mo$report, value = TRUE)
cat(tree_html)
```

Reading this top to bottom is a complete migration plan: create the two missing metadata files, move the data file into `data/` (renamed to the Psych-DS `source-*` convention) and the script into `analysis/`, and the repository will comply. The `convert_psychds()` function performs exactly these steps automatically.

## The summary counts

```{r}
mo$summary_table |>
  knitr::kable()
```

| Column | Meaning |
|---|---|
| `required_met` / `required_missing` | required Psych-DS elements present / absent |
| `recommended_met` / `recommended_missing` | recommended elements present / absent |
| `misplaced_n` | files that are not in their Psych-DS location |

## The traffic light

| Light | Meaning |
|---|---|
| **green** | the repository already follows the Psych-DS layout |
| **yellow** | some elements are missing or files are misplaced |
| **red** | required elements (like `data/` or the description file) are absent |
| **na** | there were no files to assess |

## From gap report to a compliant dataset: `convert_psychds()`

`psychds_check` tells you *what* is missing; the companion function `convert_psychds()` acts on it. Given the same repository, it writes out a Psych-DS-compliant copy — moving files into `data/`, `analysis/`, and so on, and generating the `dataset_description.json` metadata from the columns `data_check` extracted:

```{r}
#| eval: false
res <- report(paper, c("data_check", "codebook_check", "psychds_check"))
convert_psychds(res, overwrite = TRUE)
```

`convert_psychds()` can take either a paper object (it runs the checks as needed) or a **captured result** of `report(paper, ...)` — when the capture already contains `data_check` / `codebook_check` / `psychds_check`, those outputs are reused rather than re-run, and the OSF listing is not re-queried (`refresh_osf = FALSE` by default). This is why the recommended pattern is to run the checks once and hand the result to the converter. The output goes to `output_dir` (default `psychds/<paper_id>`); an existing directory is skipped unless `overwrite = TRUE`.

### What the converter does, step by step

1. **Resolves the plan.** It reuses `data_check`'s `structure` (per-file classification) and `table` (column facets), `codebook_check`'s labels and identified scales, and `psychds_check`'s placement plan (`current_path` → `target_path`).
2. **Copies every file to its Psych-DS location.** Files are resolved to their source *by row index*, not by name — a paper's several OSF components can each ship their own `demographics.csv`, and a name-keyed lookup would silently give every study the first component's bytes. A non-CSV data source (`.xlsx`/`.sav`/`.dta`/`.jasp`/…) is written out as a **real CSV** from the fully-read data frame (a byte copy would be an invalid CSV) *and* the untouched original is kept beside it, so the release carries both the reusable data and the authored artifact. Data CSVs are written **BOM-free**, because a UTF-8 BOM makes the first header read as `﻿id` and mismatches `variableMeasured`.
3. **Splits multi-study repositories.** When `data_check` assigned study groups (see [Data Check](mod-data-check.qmd)), each becomes a self-contained `study-<group>/` dataset with its own `dataset_description.json`. Files belonging to no single study (a whole-repo README/codebook, shared materials) sit at the collection root beside the study folders — following BIDS, which places shared content at the root.
4. **Writes scale definitions** as OpenScales OSD files (`scales/<code>.osd`), and **normalises trial-level data** into Behaverse `paradata/<instrument>.json`, cross-linking the two.
5. **Writes the paper's full text** into `documentation/`, and a **provenance `logs/` folder** (the file manifest, and the check results / module tables when those modules ran), each explained in the root README.

The generated dataset is validated with `psychds_validate()`, Metacheck's built-in checker for the standard, so a single-study converted output is guaranteed to pass. `convert_psychds()` returns (invisibly) a list with `output_dir`, `n_files_copied`, `n_studies`, `descriptions`, `collection`, `n_scales`, `n_paradata`, `fulltext`, `logs`, `skipped`, and `copy_failed`.

## The schemas Metacheck writes

`convert_psychds()` is where several standards come together. Understanding them explains what is in the generated files.

### Psych-DS (`dataset_description.json`)

The heart of a Psych-DS dataset is a root `dataset_description.json`: schema.org JSON-LD with `@type` `Dataset`, `schemaVersion` `"Psych-DS 1.5.1"`, the paper's `name`, `author`, `identifier` (DOI), and `keywords`, and — the substantive part — a `variableMeasured` array describing every column of every data CSV. Psych-DS requires *each* CSV column to appear here.

Each entry is a schema.org `PropertyValue` built by the converter from `data_check`'s facets and `codebook_check`'s labels. Because schema.org has no native properties for most psychometric metadata, Metacheck records those under a namespaced `metacheck:` extension whose fields map onto **DDI** concepts:

| Field | Source | DDI / schema.org concept |
|---|---|---|
| `name` | column name | variable name |
| `description` | codebook label | variable label |
| `measurementTechnique` | identified scale name | the instrument a variable was measured with |
| `metacheck:scale` | scale `{name, code, source, confidence}` | DDI variable group; `code` cross-references `scales/<code>.osd` |
| `metacheck:measurementLevel` | measurement-level facet | DDI `@classificationLevel` |
| `unitText` | unit facet | schema.org `unitText` (DDI UnitType) |
| `metacheck:concept` | concept facet | DDI Variable → Concept |
| `metacheck:role` | role facet | DDI VariableRole |
| `metacheck:codeList` | decoded value labels | DDI CodeList (one `PropertyValue` per code → label) |
| `metacheck:missingValues` | declared missing scheme | DDI MissingValues |
| `metacheck:question` / `metacheck:universe` | codebook question/filter | DDI QuestionText / Universe |
| `minValue`/`maxValue`/`metacheck:statistics` | `data_col_stats()` | schema.org numeric bounds + a stats block |
| `valuePattern` | sample values | observed category pattern |

An **empty** (all-NA) column gets a minimal stub entry naming it and flagging it empty, so the CSV column is documented without inventing statistics.

### The collection: `collection.json`

A multi-study output root is a **collection** of datasets, not itself a Psych-DS dataset, so it carries *no* root `dataset_description.json`. Instead it gets a `collection.json` — schema.org JSON-LD with `@type` `Collection` — whose `hasPart` indexes every part: each `study-<group>/` dataset (with its variable count), the root-level shared files, the paper full text, and the logs. It is deliberately **not** named `dataset_description.json`, so the Psych-DS validator (which only ever opens a file of that exact name) never validates the root — you validate each `study-*/` folder instead. This is why a multi-study convert is not run through `psychds_validate()` at the root: the root is correctly *not* a dataset.

### OpenScales OSD (`scales/<code>.osd`)

Every scale that `codebook_check` identifies is written as one OpenScales **OSD** file, flat at `scales/<code>.osd`. The `code` is a readable slug derived from the scale name (PANAS → `positive_and_negative_affect_schedule`) or, for an unnamed block, its column prefix. Each OSD records the scale's provenance at one of four trust levels (`.osd_code_and_provenance()`): `dictionary` (matched a known instrument in Metacheck's OpenScales-derived scale dictionary), `manuscript` (a named instrument from the paper), `self_generated` (an LLM-inferred construct label — *not* a recognised instrument), or `unnamed_block` (a coherent same-prefix rating block that could not be named). The `variableMeasured` entry for each of the scale's columns cross-references its OSD `code`, so a reader can jump from a variable to its instrument definition. When a scale matches a dictionary instrument, its reference items (including which are reverse-keyed) are available via the `scale_meta` / `scale_items` datasets.

### Behaverse trial data (`paradata/<instrument>.json`)

Trial-level task data (E-Prime, Inquisit, jsPsych, native Behaverse) is not a rectangular dataset — it is one file per participant per block. `convert_psychds()` normalises all of it into the **Behaverse Data Model** `trial` schema (pinned v26.0608, shipped in `inst/schema/`), writing one `paradata/<instrument>.json` per instrument with the full response data merged across participants — nothing is deleted. Each file is a `TrialData` document: an `Instrument` descriptor plus a `Response` array, where every response carries the 13 required Behaverse fields (`response_id`, `study_name`, `agent_id`, `instrument_id`, `trial_index`, `stimulus_id`, …) and any optional paradata the source recorded (`response_time`, `correct`, `stimulus_onset`, …). Each source format has its own reader that maps its columns onto this vocabulary — Inquisit's `latency` → `response_time`, jsPsych's `rt` → `response_time`, E-Prime's `<obj>.RT`, and so on (`convert_behaverse()`). The OSD scale file and the paradata file cross-reference each other on the canonical instrument id, so a questionnaire's scale definition and its raw timing data point at each other.

### DDI-Codebook 2.5 (the manifest)

The optional file manifest (`data_check`'s `manifest =` argument) is self-describing: its `provenance` block carries a `ddi_mapping` recording how each manifest field maps onto a **DDI-Codebook 2.5** element (`files[].file_name` → `fileDscr/fileTxt/fileName`, `files[].status` → `fileDscr/fileTxt/ProcStat`, and so on), alongside the metacheck/R version, platform, and production date needed to reproduce the archive.

## Building a whole data archive

The same machinery scales up from one repository to a **data release** — a browsable, archivable copy of the data behind an entire corpus of papers, each converted to Psych-DS. This is a good way to turn scattered OSF/GitHub/Zenodo repositories into a single, structured, machine-readable collection.

The recipe is: run the data pipeline over each paper with `data_check` set to fetch **all** files, then convert each to Psych-DS. Two `data_check` options (covered in the [Data Check](mod-data-check.qmd) chapter) make this practical for real-world repositories:

- `download = "all"` fetches every file, not just the readable data.
- `skip_types = "asset"` leaves out stimuli/media (which a release links to rather than hosts), and `peek_zips = TRUE` looks inside zips and only downloads those containing data.

```{r}
#| eval: false
for (paper in papers) {
  chain <- report(paper,
                  c("data_check", "codebook_check", "psychds_check"),
                  args = list(data_check = list(
                    download = "all", skip_types = "asset", peek_zips = TRUE)))
  convert_psychds(chain,  output_dir = file.path("archive", paper_id(paper), "psychds"))
  convert_codebook(chain, output_dir = file.path("archive", paper_id(paper), "codebook"))
}
```

### Zips are unpacked into the archive

When `peek_zips = TRUE`, a downloaded zip that contains data is unpacked and its inner data files are placed into the archive individually — so a `data.zip` becomes real, browsable `data/` files rather than an opaque bundle. Two helper functions expose this directly:

- **`zip_peek(url)`** reads a remote zip's file listing (names and sizes) via an HTTP range request, *without* downloading it.
- **`zip_decision(url)`** peeks and classifies the contents, returning whether the zip is worth downloading (contains data) or should just be linked (only stimuli).

### An archive-level catalogue

A release of many datasets needs a top-level description tying them together. **`data_catalog()`** scans a built archive directory and writes a `catalog.json` — a schema.org [`DataCatalog`](https://schema.org/DataCatalog) listing every dataset with its title, authors, DOI, and file count — plus a `catalog.csv` for quick inspection. This is the object that makes the release a single discoverable archive (and is harvested by Google Dataset Search), rather than a set of unconnected folders.

```{r}
#| eval: false
data_catalog("archive", papers = papers, name = "My data release")
```

::: callout-tip
For a large corpus, first run `data_check` with `download = "none"` and `manifest = "..."` to write a per-paper file manifest (file names, sizes, types) *without* downloading anything. Inspecting those manifests lets you choose the size caps knowing exactly which files each value would include or exclude, before committing to a large download.
:::

## What you get back

| Element | What it contains |
|---|---|
| `$traffic_light` | `green` / `yellow` / `red` / `na` |
| `$summary_text` | plain-text compliance summary |
| `$summary_table` | the required/recommended/misplaced counts above |
| `$table` | one row per file: `file_name`, `data_type`, `current_path`, `target_path`, `status` |
| `$report` | formatted report, including the target directory tree with missing pieces marked |

## Related

- [Data Check](mod-data-check.qmd) classifies the files that `psychds_check` re-organises.
- [Codebook Check](mod-codebook-check.qmd) checks variable *documentation*, which becomes the `variableMeasured` metadata in a Psych-DS dataset.
- See the [OSF Prereg Schemas](osf-schemas.qmd) appendix for related standards Metacheck understands.
