---
title: "Preregistration Check"
vignette: >
  %\VignetteIndexEntry{Preregistration Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## The basics

A preregistration is a document, written and time-stamped before a study is run, that states what the researchers plan to do — their hypotheses, sample size, and analysis plan. `prereg_check` looks for links to preregistrations hosted on the OSF (Open Science Framework) or AsPredicted, two websites researchers commonly use for this, and reads out their contents.

Here is the module running on a demo paper, which links to two preregistrations:

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "prereg_check")
```

Running a module always works the same way: `module_run(paper, "module_name")`. The printed result above is a short report: a coloured light and a one-line summary.

If you want to work with the result instead of just reading it, save it to a variable and look inside it:

```{r}
mo <- module_run(paper, "prereg_check")
mo$table[, c("template_name", "title", "id", "sample_size")] |>
  knitr::kable()
```

`mo$table` has one row per preregistration found, with its contents organised into columns such as the sample-size plan shown above. A green light only means the module found a preregistration and could read its contents — it does not mean the study was actually run the way the preregistration describes. Always read the preregistration's own fields yourself and compare them to what the manuscript reports.

## What it checks

The `prereg_check` module finds links to **preregistrations** on the OSF and AsPredicted in the manuscript, retrieves the preregistration text, and organises its contents into a single standardised template. This makes it much easier for a reviewer to compare what was preregistered against what the manuscript reports.

::: callout-note
This module makes **live network calls** to the OSF and AsPredicted to retrieve preregistration content. You need an internet connection to run the code below. (When this book is built, these calls run against the live services.)
:::

## Running the module

`demopaper()` links to two preregistrations, so it is a good example.

```{r}
paper <- demopaper()
mo <- module_run(paper, "prereg_check")
mo$traffic_light
mo$summary_text
```

The `table` has one row per preregistration, with the template fields organised into columns. Here are a few of them:

```{r}
mo$table[, c("template_name", "title", "id", "sample_size")] |>
  knitr::kable()
```

The full set of fields the module standardises includes the study design, hypotheses, sample-size rationale, stopping rule, variables, statistical tests, inference criteria, and exclusion criteria:

```{r}
names(mo$table)
```

## Reading a single preregistration's contents

Because each preregistration becomes a row, you can inspect one in detail by transposing it:

```{r}
prereg <- mo$table[1, c("template_name", "sample_size", "statistical_tests", "data_exclusion_criteria")]
t(prereg) |>
  knitr::kable()
```

## Options and limitations

`prereg_check` takes only the `paper` argument. It **cannot** extract information from:

- unstructured preregistration templates (where everything is in one free-text field), or
- preregistrations stored as uploaded text documents on the OSF rather than in the structured form.

For the details of which OSF registration templates exist and how they are identified, see the [OSF Registration Schemas](osf-schemas.qmd) chapter.

## Notes

If you want Metacheck to support additional preregistration platforms, reach out to the development team.
