---
title: "Code Check"
vignette: >
  %\VignetteIndexEntry{Code Check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `code_check` module analyses the **code files** shared with a paper. Building on the file list from [repo_check](mod-repo-check.qmd), it identifies code files (R, SAS, SPSS, Stata), then checks each for signs of reproducibility problems: whether files parse without errors, whether they contain comments, whether they use **absolute file paths** (which break on other machines), and whether files they try to load are actually present in the repository.

::: callout-note
This module makes **live network calls** to retrieve and inspect code files from the linked repositories. You need an internet connection to run the code below.
:::

## Running the module

```{r}
paper <- demopaper()
mo <- module_run(paper, "code_check")
mo$traffic_light
cat(mo$summary_text)
```

The `table` has one row per code file, with the reproducibility checks as columns:

```{r}
mo$table[, c("file_name", "language", "parse_error", "absolute_paths",
             "percentage_comment", "loaded_files_missing")] |>
  knitr::kable()
```

## A clean example and one with problems

Common problems the module surfaces:

- **Absolute paths** (`absolute_paths == TRUE`) — e.g. `read.csv("C:/Users/me/data.csv")` — which will not run on anyone else's computer.
- **Missing loaded files** (`loaded_files_missing > 0`) — the script reads a file that was not shared.
- **Parse errors** (`parse_error == TRUE`) — the file is not valid code.

```{r}
mo$table[, c("file_name", "absolute_paths", "loaded_files_missing", "parse_error")] |>
  knitr::kable()
```

## Options

`code_check` shares `repo_check`'s local-file options, plus a limit on how many files to inspect:

```{r}
#| eval: false
# inspect at most 5 files (default is 20)
module_run(paper, "code_check", file_limit = 5)

# check code in a local folder only, no online lookups
module_run(paper, "code_check", local_path = "path/to/files", local_only = TRUE)
```

See the [Local Files](local-files.qmd) chapter for a full walkthrough of checking local code, including cloud-synced folders.

## Notes

The module reports on *potential* problems. An absolute path or a missing file is not always an error — but it is usually worth a second look if you want the analysis to be reproducible.
