---
title: "PubPeer Comments"
vignette: >
  %\VignetteIndexEntry{PubPeer Comments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `ref_pubpeer` module cross-references the paper's **cited references** against [PubPeer](https://pubpeer.com/), a platform for post-publication peer review. If a paper you cite has comments on PubPeer, that may be worth knowing — comments can range from minor clarifications to serious concerns about the work.

The module reports which cited references have PubPeer comments, how many, and a link, so you can read the discussion and judge its relevance yourself.

Statcheck comments on PubPeer are ignored. Statcheck automatically posted comments on [tens of thousands of psychology articles](https://retractionwatch.com/2016/09/02/heres-why-more-than-50000-psychology-studies-are-about-to-have-pubpeer-entries/), so including them would flag almost every cited paper and drown out substantive human comments.

::: callout-note
This module makes **live network calls** to PubPeer (matching references by DOI). You need an internet connection to run the code below.
:::

## Running the module

`demopaper()` cites a reference that has PubPeer comments.

```{r}
paper <- demopaper()
mo <- module_run(paper, "ref_pubpeer")
mo$traffic_light
mo$summary_text
```

The `table` lists each matched reference, the number of comments, and a link:

```{r}
mo$table[, c("doi", "total_comments", "url")] |>
  knitr::kable()
```

## Running on many papers

```{r}
#| eval: false
mo <- module_run(psychsci[1:10], "ref_pubpeer")
mo$summary_table
```

## Interpreting the result

A `green`/empty result means none of the cited references had PubPeer comments. A flagged result does **not** mean a citation is bad — it means there is a discussion you may want to read. Follow the `url` to the PubPeer thread and judge the comments in context.

## Options

`ref_pubpeer` takes only the `paper` argument.

## Notes

In the Shiny app, querying PubPeer is an opt-in checkbox, because it sends the cited DOIs to an external service.
