---
title: "Non-Significant p-Values (stat_p_nonsig)"
vignette: >
  %\VignetteIndexEntry{Non-Significant p-Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
library(metacheck)
```

## What it checks

The `stat_p_nonsig` module finds every **non-significant p-value** (p ≥ .05) so a reader can check whether the surrounding text interprets it correctly. A common error is to treat a non-significant result as evidence that there is *no effect* ("there was no difference between groups, p = .27"), when a non-significant result is simply inconclusive.

The module makes no judgement itself — it only surfaces the sentences so you can read how each result is described.

This module is **fully offline**.

## Running the module

```{r}
#| results: asis
paper <- demopaper()
module_run(paper, "stat_p_nonsig")
```

The `table` lists each non-significant p-value with its comparison and value:

```{r}
mo <- module_run(paper, "stat_p_nonsig")
mo$traffic_light
mo$summary_text
mo$table[, c("text", "p_comp", "p_value", "significance")] |>
  knitr::kable()
```

## Running on many papers

```{r}
mo <- module_run(psychsci[1:20], "stat_p_nonsig")
head(mo$summary_table) |>
  knitr::kable()
```

## A clean example and one with problems

A result that is correctly described as inconclusive is fine; one that claims "no effect" from a non-significant test is the kind of thing to check. The module flags both, and you read the sentence:

```{r}
flagged <- test_paper("There was no difference between the groups, t(40) = 1.10, p = .28.")
mo <- module_run(flagged, "stat_p_nonsig")
mo$table[, c("text", "p_value")] |>
  knitr::kable()
```

A paper reporting only significant results returns nothing to check:

```{r}
clean <- test_paper("The effect was significant, t(28) = 2.40, p = .02.")
module_run(clean, "stat_p_nonsig")$traffic_light
```

## Options

`stat_p_nonsig` takes only the `paper` argument.

## Validation

> In a sample of 194 papers with 1602 instances of non-significant p-values, this module correctly detected 1486 of them and incorrectly identified 153. 91% of detections were true instances (positive predictive value).
