---
title: "Reading in a Paper"
vignette: >
  %\VignetteIndexEntry{Reading in a Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
library(metacheck)
```

The first step to using Metacheck is to read in a manuscript. Currently, Metacheck relies on GROBID, which works with PDF files. If you have a manuscript in a different format, save a copy as a PDF. You have two options to convert a PDF to the GROBID .xml format. First, you can use a local version of GROBID. We will explain this below, but it requires more technical knowledge, and sufficient space on your harddrive. Alternatively, you can use an online server. As this is much simpler, and you can use a GROBID server at Eindhoven University of Technology that is GDPRR compliant and does not store any data. This is the approach we will use by default in this manual. Nevertheless, you might prefer not to send your manuscript to an external server, or you might not be allowed to. We have developed Metacheck to always work locally, without an internet connection, to also make it available in these more restrictive situations. 

The function `convert()` can read PDF files and save them in [JSON format](https://www.scienceverse.org/schema/paper.json). This requires an internet connection and takes a few seconds per paper. The conversion process needs to be done only once and the converted file can be saved for later use. If you don't supply an `api_url`, Metacheck will use a [list of active servers](https://www.scienceverse.org/metacheck/convert.json) and choose the first available. By default, this is a GROBID server at Eindhoven University of Technology that is GDPR compliant.

You can also supplement the reference section created by GROBID with information retrieved from CrossRef by setting `crossref_lookup = TRUE`. This can increase the accuracy of the information in the reference list. Retrieving this information also requires an internet connection, and can take some time. You can always add this information later with the `add_bibmatch()` function.

```{r, eval = FALSE}
pdf_file <- demofile("pdf")
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     crossref_lookup = TRUE)
```

```{r, include = FALSE}
# doesn't require a call to the grobid server
json_file <- demofile("json")
```

## Local conversion through GROBID in Docker

You can set up your own local GROBID server following instructions from <https://grobid.readthedocs.io/>. The easiest way is to use Docker. The following code installs GROBID 0.9.0 if you run it in your system terminal.

``` bash
docker run --rm --init --ulimit core=0 -p 8070:8070 lfoppiano/grobid:0.9.0
```

If you set the `api_url` to the local path <http://localhost:8070> (as illustrated below) the `convert()` function will default to the local GROBID server. Metacheck will by default use the local GROBID server if it is automatically detected.

```{r, eval = FALSE}
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     method = "grobid",
                     api_url = "http://localhost:8070")
```

## Load from JSON or XML

The function `read()` reads in papers created by `convert()`. It accepts both the **bibr JSON** files and the **GROBID XML** (TEI) files that conversion produces, so you can use whichever you have.

```{r}
# read a converted JSON file
paper <- read(json_file)
```

```{r}
# read a GROBID XML file directly
paper_xml <- read(demofile("xml"))
```

## Batch Processing

The functions `convert()` and `read()` also work on a folder of files, or a vector of paths, returning a list of JSON file paths or paper objects, respectively. Batch processing is especially useful for scientists who want to be able to examine a large number of manuscripts. For example, a metascientist might be interesting in examining the uptake of open science practices over time, and run Metacheck across all articles published in a scientific journal. As part of Metacheck, we have created a paper repository of more than 15.000 open access articles that you can read in through the function `papers_load()`. See [Using the Paper Database Corpora](papers-databases.qmd) for how to load the available corpora.

When a folder contains both a `.json` and a `.xml` file for the same paper, `read()` uses the JSON and skips the duplicate XML.

## The paper object

`read()` returns a **paper object**: a structured representation of the manuscript, stored as a list of tables. Everything Metacheck does — searching text, running modules, checking references — operates on this object, so it is worth understanding what is inside it and how to get at each piece.

A paper object has the class `scivrs_paper`, and you access its parts with the `$` operator. Here are all the components:

```{r}
class(paper)
names(paper)
```

The components are summarised below; the rest of this section shows each one. Most are data frames, so you can work with them using base R or the tidyverse.

| Component | What it holds |
|---|---|
| `paper_id` | A single string identifying the paper |
| `info` | One row of paper-level metadata (title, DOI, type, ...) |
| `author` | One row per author |
| `text` | One row per sentence |
| `section` | One row per heading/section |
| `url` | Links found in the text |
| `bib` | The reference list |
| `xref` | Cross-references linking sentences to bib entries, tables, and figures |
| `figure` | Figures |
| `table` | Tables |
| `eq` | Extracted statistics and equations |
| `bib_match` | Reference metadata matched from CrossRef and other services |

**paper_id**

A single string identifying the paper (the user-supplied id, the DOI, or the filename):

```{r}
paper$paper_id
```

**info**

A one-row data frame of paper-level metadata. Use `t()` to read it as a column:

```{r}
t(paper$info) |>
  knitr::kable(col.names = "value")
```

Pull out a single field directly:

```{r}
paper$info$title
paper$info$doi
```

**author**

One row per author, in paper order:

```{r}
paper$author[, c("given", "family", "affiliation", "corresponding")] |>
  knitr::kable()
```

**text**

One row per sentence (excluding the title, headings, and references). This is the table that `text_search()` searches.

```{r}
nrow(paper$text)
head(paper$text[, c("text", "section_id", "paragraph_id")]) |>
  knitr::kable()
```

**section**

One row per heading. `section_type` is Metacheck's best guess at the kind of section, which modules use to focus on (say) only the results:

```{r}
paper$section[, c("section_id", "header", "section_type")] |>
  knitr::kable()
```

**url**

Every external link found in the text, with the sentence (`text_id`) it appeared in:

```{r}
paper$url |>
  knitr::kable()
```

**bib**

The reference list, parsed into columns:

```{r}
paper$bib[, c("bib_id", "authors", "year", "title")] |>
  knitr::kable()
```

**xref**

Cross-references that link a sentence (`text_id`) to an item it cites — a bib entry, table, or figure (`xref_type`):

```{r}
paper$xref |>
  knitr::kable()
```

**figure and table**

Figures and tables each get a row, linked to the `section` table by `section_id`:

```{r}
paper$figure[, c("figure_id", "section_id", "page_number")] |>
  knitr::kable()
paper$table[, c("table_id", "section_id", "page_number")] |>
  knitr::kable()
```

**eq**

Statistics and equations extracted from the text, broken into a left-hand side, degrees of freedom, comparator, and right-hand side. Many statistics modules read from this table rather than re-scanning the text:

```{r}
head(paper$eq) |>
  knitr::kable()
```

**bib_match**

When you convert with `crossref_lookup = TRUE` (or run a reference module), Metacheck matches each reference against services like CrossRef and stores the retrieved metadata here, keyed by `bib_id`:

```{r}
paper$bib_match[, c("bib_id", "service", "doi", "title")] |>
  knitr::kable()
```

## Working with a list of papers

A *paperlist* is simply a list of paper objects — for example the built-in `psychsci` dataset (250 open-access *Psychological Science* articles). To pull the same component from every paper into one combined table, use `paper_table()`:

```{r}
paper_table(psychsci[1:5], "info", c("title", "doi")) |>
  knitr::kable()
```

Most functions and every module accept either a single paper or a paperlist, so the same code scales from one paper to a whole corpus. See [Using the Paper Database Corpora](papers-databases.qmd) for the available corpora.
