---
title: "Reading in a Paper"
vignette: >
  %\VignetteIndexEntry{Reading in a Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
library(metacheck)
```

The first step to using Metacheck is to read in a manuscript. Currently, Metacheck relies on GROBID, which works with PDF files. If you have a manuscript in a different format, save a copy as a PDF. You have two options to convert a PDF to the GROBID .xml format. First, you can use a local version of GROBID. We will explain this below, but it requires more technical knowledge, and sufficient space on your harddrive. Alternatively, you can use an online server. As this is much simpler, and you can use a GROBID server at Eindhoven University of Technology that is GDPRR compliant and does not store any data. This is the approach we will use by default in this manual. Nevertheless, you might prefer not to send your manuscript to an external server, or you might not be allowed to. We have developed Metacheck to always work locally, without an internet connection, to also make it available in these more restrictive situations. 

The function `convert()` can read PDF files and save them in [JSON format](https://www.scienceverse.org/schema/paper.json). This requires an internet connection and takes a few seconds per paper. The conversion process needs to be done only once and the converted file can be saved for later use. If you don't supply an `api_url`, Metacheck will use a [list of active servers](https://www.scienceverse.org/metacheck/convert.json) and choose the first available. By default, this is a GROBID server at Eindhoven University of Technology that is GDPR compliant.

You can also supplement the reference section created by GROBID with information retrieved from CrossRef by setting `crossref_lookup = TRUE`. This can increase the accuracy of the information in the reference list. Retrieving this information also requires an internet connection, and can take some time. You can always add this information later with the `add_bibmatch()` function.

```{r, eval = FALSE}
pdf_file <- demofile("pdf")
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     crossref_lookup = TRUE)
```

```{r, include = FALSE}
# doesn't require a call to the grobid server
json_file <- demofile("json")
```

### Local conversion through GROBID in Docker

You can set up your own local GROBID server following instructions from <https://grobid.readthedocs.io/>. The easiest way is to use Docker. The following code installs GROBID 0.9.0 if you run it in your system terminal.

``` bash
docker run --rm --init --ulimit core=0 -p 8070:8070 lfoppiano/grobid:0.9.0
```

If you set the `api_url` to the local path <http://localhost:8070> (as illustrated below) the `convert()` function will default to the local GROBID server. Metacheck will by default use the local GROBID server if it is automatically detected.

```{r, eval = FALSE}
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     method = "grobid",
                     api_url = "http://localhost:8070")
```

### Load from JSON or XML

The function `read()` reads in papers created by `convert()`. It accepts both the **bibr JSON** files and the **GROBID XML** (TEI) files that conversion produces, so you can use whichever you have.

```{r}
# read a converted JSON file
paper <- read(json_file)
```

```{r}
# read a GROBID XML file directly
paper_xml <- read(demofile("xml"))
```

### Batch Processing

The functions `convert()` and `read()` also work on a folder of files, or a vector of paths, returning a list of JSON file paths or paper objects, respectively. Batch processing is especially useful for scientists who want to be able to examine a large number of manuscripts. For example, a metascientist might be interesting in examining the uptake of open science practices over time, and run Metacheck across all articles published in a scientific journal. As part of Metacheck, we have created a paper repository of more than 15.000 open access articles that you can read in through the function `papers_load()`. See [Using the Paper Database Corpora](papers-databases.qmd) for how to load the available corpora.

When a folder contains both a `.json` and a `.xml` file for the same paper, `read()` uses the JSON and skips the duplicate XML.

## Paper Components

Paper objects contain structured information retrieved from the converted document. This in, including info, references, and citations.

### Info

```{r}
paper$info
```


### References

The references are provided in a tabular format.

```{r, eval = FALSE}
paper$bib
```

```{r, echo = FALSE}
isna <- paper$bib |> lapply(is.na) |> sapply(all)
knitr::kable(paper$bib[, !isna])
```

### Cross-References

Cross-references are also provided in a tabular format, with `xref_id` to match the bibliography table.

```{r, eval = FALSE}
paper$xref
```

```{r, echo = FALSE}
knitr::kable(paper$xref)
```

### Batch

The `psychsci` built-in dataset contains all 250 open-access articles published in *Psychological Science* from 2004 to 2014. *Psychological Science* is the flagship journal of the Association for Psychological Science and publishes empirical research from all subfields of psychology. The PDFs were converted to metacheck paper objects using `convert()` and stored in the package. This dataset is used throughout the documentation to demonstrate how to work with a list of papers.

There are functions to combine the information from a list of papers, like `psychsci`.

```{r}
paper_table(psychsci[1:5], "info", c("title", "doi"))
```

```{r}
paper_table(psychsci[1:5], "bib") |>
  dplyr::filter(!is.na(doi), doi != "")
```


## Next steps

Once you have a paper object, you can search its text directly (see [Text Search](text-search.qmd)) or run checks on it. metacheck comes with a set of built-in modules:

```{r}
module_list()
```

Each module is documented in its own chapter in the [Modules](modules.qmd) part. To learn how the module system works and how to write your own, see [Modules](modules.qmd) and [Creating Modules](creating-modules.qmd).
