---
title: "Reading in a Paper"
vignette: >
  %\VignetteIndexEntry{Reading in a Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
library(metacheck)
```

The function `convert()` can read PDF files and save them in [JSON format](https://www.scienceverse.org/schema/paper.json). This requires an internet connection and takes a few seconds per paper. The conversion process needs to be done only once and the converted file can be saved for later use. If you don't supply an `api_url`, Metacheck will use a [list of active servers](https://www.scienceverse.org/metacheck/convert.json) and choose the first available. By default, this is a GROBID server at Eindhoven University of Technology tht meets GDPR requirements.

You can also supplement the reference section created by GROBID with information retrieved from Crossref. This can increase the accuracy of the information in the reference list. Retrieving this information also requires an internet connection, an can take some time. You can always add this information later with the `add_bibmatch()` function.

```{r, eval = FALSE}
pdf_file <- demofile("pdf")
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     crossref_lookup = TRUE)
```

```{r, include = FALSE}
# doesn't require a call to the grobid server
json_file <- demofile("json")
```


You can set up your own local grobid server following instructions from <https://grobid.readthedocs.io/>. The easiest way is to use Docker. The following code installs GROBID 0.9.0.

``` bash
docker run --rm --init --ulimit core=0 -p 8070:8070 lfoppiano/grobid:0.9.0
```

If you set the `api_url` to the local path <http://localhost:8070> the `convert()` function will default to the local GRBOID server. Metacheck will also by default use the local GROBID server if it is automatically detected.

```{r, eval = FALSE}
json_file <- convert(file_path = pdf_file,
                     save_path = "converted",
                     method = "grobid",
                     api_url = "http://localhost:8070")
```


### Load from JSON

The function `read()` can read converted JSON files created by the `convert` function.

```{r}
paper <- read(json_file)
```


### Load from non-PDF document

To take advantage of GROBID's ability to parse references and other aspects of papers, for now the best way is to convert your papers to PDF. We are developing our own custom backend called `bibr` which will also be able to convert DOC and DOCX files directly.

### Batch Processing

The functions `convert()` and `read()` also work on a folder of files, or a vector of paths, returning a list of JSON file paths or paper objects, respectively.

## Paper Components

Paper objects contain structured information retrieved from the converted document. This in, including info, references, and citations.

### Info

```{r}
paper$info
```


### References

The references are provided in a tabular format.

```{r, eval = FALSE}
paper$bib
```

```{r, echo = FALSE}
isna <- paper$bib |> lapply(is.na) |> sapply(all)
knitr::kable(paper$bib[, !isna])
```

### Cross-References

Cross-references are also provided in a tabular format, with `xref_id` to match the bibliography table.

```{r, eval = FALSE}
paper$xref
```

```{r, echo = FALSE}
knitr::kable(paper$xref)
```

### Batch

The `psychsci` built-in dataset contains all 250 open-access articles published in *Psychological Science* from 2004 to 2014. *Psychological Science* is the flagship journal of the Association for Psychological Science and publishes empirical research from all subfields of psychology. The PDFs were converted to metacheck paper objects using `convert()` and stored in the package. This dataset is used throughout the documentation to demonstrate how to work with a list of papers.

There are functions to combine the information from a list of papers, like `psychsci`.

```{r}
paper_table(psychsci[1:5], "info", c("title", "doi"))
```

```{r}
paper_table(psychsci[1:5], "bib") |>
  dplyr::filter(!is.na(doi), doi != "")
```


## Search Text

You can access a parsed table of the full text of the paper via `paper$text`, but you may find it more convenient to use the function `text_search()`. The defaults return a data table of each sentence, with the section type, header, div, paragraph and sentence numbers, and file name. (The section type is a best guess from the headers, so may not always be accurate.)

```{r}
text <- text_search(paper)
```

```{r, echo = FALSE}
head(text) |>
  knitr::kable()
```


### Pattern

You can search for a specific word or phrase by setting the `pattern` argument. The pattern is a regex string by default; set `fixed = TRUE` if you want to find exact text matches.

```{r}
text <- text_search(paper, pattern = "metacheck")
```

```{r, echo = FALSE}
knitr::kable(text)
```


### Return

Set `return` to one of "sentence", "paragraph", "section", or "match" to control what gets returned.

```{r}
text <- text_search(paper, "GitHub",
                    return = "paragraph")
```

```{r, echo = FALSE}
knitr::kable(text)
```



### Regex matches

You can also return just the matched text from a regex search by setting `return = "match"`. The extra `...` arguments in `text_search()` are passed to `grep()`, so `perl = TRUE` allows you to use more complex regex, like below.

```{r}
pattern <- "[a-zA-Z]\\S*\\s*(=|<)\\s*[0-9\\.,-]*\\d"
text <- text_search(paper, pattern, return = "match", perl = TRUE)
```

```{r, echo = FALSE}
knitr::kable(text)
```


### Expand Text

You can expand the text returned by `text_search()` or a module with `text_expand()`.

```{r}
marginal <- text_search(paper, "marginal") |>
  text_expand(paper, plus = 1, minus = 1)

marginal[, c("text", "expanded")]
```


## Repository Functions

Metacheck can find links to research repositories in a paper, retrieve the list of files they contain, and download those files for further checking. Four online services are supported, as well as local folders on your own computer.

| Repository | Link function | Info / file list | Download |
|---|---|---|---|
| OSF | `osf_links()` | `osf_info()` | `osf_file_download()` |
| GitHub | `github_links()` | `github_files()`, `github_info()` | — |
| ResearchBox | `rbox_links()` | `rbox_info()` | `rbox_file_download()` |
| Zenodo | `zenodo_links()` | `zenodo_info()` | `zenodo_file_download()` |
| Local folder | — | `local_files()` | — |

The `repo_check` and `code_check` modules use all of these automatically: `repo_check` finds all repository links in a paper and builds a unified file list, and `code_check` then analyses any code files in that list. See the [GitHub](github.html) and [Local Files](local-files.html) articles for more detail on those two sources.

### OSF Links and IDs

Get any OSF links from a paper or list of papers.

```{r}
links <- osf_links(psychsci)

links$href |> unique() |> head()
```


You can see that some of them have rogue spaces or view-only links. The function `osf_check_id()` takes most formats of OSF links (with or without https:// and osf.io/, as well as the 25-character waterbutler IDs) and converts them to short IDs. 

```{r}
osf_ids <- osf_check_id(links$href) |> unique()

head(osf_ids)
```


However, all of the `osf_***()` functions fix IDs for you and handle duplicate IDs without making extra API calls, so you don't need to add this step to most workflows.

### OSF Info

Get basic information about OSF links, such as the name, description, osf_type (nodes, files, preprints, registrations, users, set to "private" if you don't have authorisation to view it, and "invalid" if the ), whether it is public

```{r}
#| eval: false
info <- osf_info(links[1:6, "href"])

info[, c("href","osf_id", "osf_type", "public", "category")]
```

```{r}
#| echo: false
info <- structure(list(href = c("https://osf.io/e2aks/", "https://osf.io/tvyxz/wiki/view/", 
"https://osf.io/tvyxz/wiki/view/", "https://osf.io/t9j8e/?view_only=f171281f212f4435917b16a9e581a73b", 
"https://osf.io/tvyxz/wiki/1.%20View%20the%20Badges/", "https://osf.io/eky4s/"
), osf_id = c("e2aks", "tvyxz", "tvyxz", "t9j8e?view_only=f171281f212f4435917b16a9e581a73b", 
"tvyxz", "eky4s"), name = c("Action-specific disruption of perceptual confidence", 
"Badges to Acknowledge Open Practices", "Badges to Acknowledge Open Practices", 
"PSICI-14-1362 - Failing to forget: Inhibitory control deficits compromise memory suppression in post-traumatic stress disorder", 
"Badges to Acknowledge Open Practices", "Wearing a Bicycle Helmet Can Increase Risk Taking and Sensation Seeking in Adults"
), description = c("Data and analysis scripts for:\nFleming, S.M., Manisaclco, B., Ko, Y., Amendi, N., Ro, T. &amp; Lau, H. (2015) Action-specific disruption of perceptual confidence. Psychological Science https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4361353/", 
"The aim is to specify a standard by which we can say that a scientific study has been conducted in accordance with open-science principles and provide visual icons to allow advertising of such good behaviours.", 
"The aim is to specify a standard by which we can say that a scientific study has been conducted in accordance with open-science principles and provide visual icons to allow advertising of such good behaviours.", 
"Open materials for Psychological Science paper: Failing to forget: Inhibitory control deficits compromise memory suppression in post-traumatic stress disorder", 
"The aim is to specify a standard by which we can say that a scientific study has been conducted in accordance with open-science principles and provide visual icons to allow advertising of such good behaviours.", 
""), osf_type = c("nodes", "nodes", "nodes", "nodes", "nodes", 
"nodes"), public = c(TRUE, TRUE, TRUE, FALSE, TRUE, TRUE), category = c("project", 
"project", "project", "project", "project", "project"), registration = c(FALSE, 
FALSE, FALSE, FALSE, FALSE, FALSE), preprint = c(FALSE, FALSE, 
FALSE, FALSE, FALSE, FALSE), self = c("https://api.osf.io/v2/nodes/e2aks/", 
"https://api.osf.io/v2/nodes/tvyxz/", "https://api.osf.io/v2/nodes/tvyxz/", 
"https://api.osf.io/v2/nodes/t9j8e/?view_only=f171281f212f4435917b16a9e581a73b", 
"https://api.osf.io/v2/nodes/tvyxz/", "https://api.osf.io/v2/nodes/eky4s/"
), children = c("https://api.osf.io/v2/nodes/e2aks/children/", 
"https://api.osf.io/v2/nodes/tvyxz/children/", "https://api.osf.io/v2/nodes/tvyxz/children/", 
"https://api.osf.io/v2/nodes/t9j8e/children/?view_only=f171281f212f4435917b16a9e581a73b", 
"https://api.osf.io/v2/nodes/tvyxz/children/", "https://api.osf.io/v2/nodes/eky4s/children/"
), files = c("https://api.osf.io/v2/nodes/e2aks/files/", "https://api.osf.io/v2/nodes/tvyxz/files/", 
"https://api.osf.io/v2/nodes/tvyxz/files/", "https://api.osf.io/v2/nodes/t9j8e/files/?view_only=f171281f212f4435917b16a9e581a73b", 
"https://api.osf.io/v2/nodes/tvyxz/files/", "https://api.osf.io/v2/nodes/eky4s/files/"
), parent = c(NA_character_, NA_character_, NA_character_, NA_character_, 
NA_character_, NA_character_), project = c("e2aks", "tvyxz", 
"tvyxz", "t9j8e", "tvyxz", "eky4s")), row.names = c(NA, -6L), class = c("tbl_df", 
"tbl", "data.frame"))

knitr::kable(info[, c("href","osf_id", "osf_type", "public", "category")])
```



View-only links may be listed in the table as public = FALSE if they haven't been made publicly available. This means they are not searchable and only accessible with the view_only link.

You can set the argument `recursive = TRUE` to also retrieve information about all nodes and files that are contained by the OSF link.

```{r}
#| eval: false
all_contents <- osf_info(links$href[1], recursive = TRUE)

all_contents[, c("osf_id", "name")]
```


```{r}
#| echo: false
structure(list(
  osf_id = c("e2aks", "7jh5v", "pj4e8", "553e66b48c5e4a219919e0e7", 
"553e58658c5e4a219919a627", "553e7e168c5e4a21991a4dab", "553e58658c5e4a219919a628", 
"553e58658c5e4a219919a629", "553e58658c5e4a219919a62a", "553e58658c5e4a219919a62b", 
"553e58658c5e4a219919a62c", "553e7e168c5e4a21991a4dac"), 
  name = c("Action-specific disruption of perceptual confidence", 
"Data", "Analysis scripts", "osfstorage", "osfstorage", "osfstorage", 
"Mratio_all.txt", "allData_orientation.txt", "allData_contrast_M1.txt", 
"allData_contrast_PMC.txt", "Mratio_contrast_M1.txt", "tms_analysis.R"
)), 
  class = "data.frame", row.names = c(NA, -12L)) |>
  knitr::kable()
```


### Download OSF Files

OSF projects let you organise information into nested components, and files within those components. Therefore, to retrieve all of the files associate with a project, you may need to navigate to several components and download zip files for the files from each components, then reorganise and rename the downloaded folders. 

The function `osf_file_download()` does all of this for you, recreating a folder structure based on the component names and downloading all files smaller than `max_file_size` (defaults to 10 MB) up to a total size of `max_download_size` (defaults to 100 MB).

```{r}
#| eval: false
osf_file_download(osf_id = "pngda",
                  download_to = ".", 
                  max_file_size = 1, 
                  max_download_size = 10)
```

```
Starting retrieval for pngda
- omitting metacheck.png (1.5MB)
Downloading files [=====================] 24/24 00:00:35
```

```{r}
#| eval: false
list.files("pngda", recursive = TRUE)
```

```{r}
#| echo: false

# list.files("articles/pngda/", recursive = TRUE) |> dput()
c(
  "Data/Individual/data-01.csv",
  "Data/Individual/data-02.csv",
  "Data/Individual/data-03.csv",
  "Data/Individual/data-04.csv",
  "Data/Individual/data-05.csv",
  "Data/Individual/data-06.csv",
  "Data/Individual/data-07.csv",
  "Data/Individual/data-08.csv",
  "Data/Individual/data-09.csv",
  "Data/Individual/data-10.csv",
  "Data/Individual/data-11.csv",
  "Data/Individual/data-12.csv",
  "Data/Individual/data-13.csv",
  "Data/Individual/data-14.csv",
  "Data/Processed Data/processed-data.csv",
  "Data/Raw Data/data.xlsx",
  "Data/Raw Data/nest-1/nest-2/nest-3/nest-4/test-4.txt",
  "Data/Raw Data/nest-1/nest-2/nest-3/test-3.txt",
  "Data/Raw Data/nest-1/nest-2/test-2.txt",
  "Data/Raw Data/nest-1/README",
  "Data/Raw Data/nest-1/test-1.txt",
  "Data/Raw Data/README",
  "README"
)
```




### GitHub, ResearchBox, and Zenodo

The same pattern — find links, retrieve file lists, optionally download — applies to the other three services.

```{r}
#| eval: false
# GitHub
gh_links  <- github_links(paper)
gh_files  <- github_files(gh_links$href, recursive = TRUE)

# ResearchBox
rb_links  <- rbox_links(paper)
rb_info   <- rbox_info(rb_links)

# Zenodo
z_links   <- zenodo_links(paper)
z_info    <- zenodo_info(z_links)
```

See the [GitHub](github.html) article for a detailed walkthrough of the GitHub functions.

### Local files

If files are not in an online repository — for example because you downloaded a zip archive from a reviewer submission, or because the authors used a service not yet supported — you can point metacheck at a local folder instead.

```{r}
#| eval: false
result <- module_run(test_paper(), "code_check",
                     local_path = "path/to/downloaded/files")
```

See the [Local Files](local-files.html) article for full details, including how to handle cloud-synced folders and multiple paths at once.

## Modules

metacheck is designed modularly, so you can add modules to check for anything. It comes with a set of pre-defined modules, and we hope people will share more modules.

### Module List

You can see the list of built-in modules with the function below.

```{r}
module_list()
```


### Running modules

To run a built-in module on a paper, you can reference it by name.

```{r}
p <- module_run(paper, "all_p_values")
```

```{r, echo = FALSE}
knitr::kable(p$table)
```


### Creating modules

You can create your own modules using R code. Modules can also contain instructions for reporting, to give "traffic lights" for whether a check passed or failed, and to include appropriate text feedback in a report. See the [modules vignette](modules.html) for more details. 


## Reports

You can generate a report from any set of modules. Check the function help for the default set.

```{r, eval = FALSE}
report(paper, output_format = "qmd")
```


See the [example report](../report-example.html).
