---
title: "Getting API Tokens for Other Repositories"
vignette: >
  %\VignetteIndexEntry{Getting API Tokens for Other Repositories}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
execute:
  eval: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Metacheck can read files directly from several online repositories: the Open Science Framework (OSF), Zenodo, Dataverse, Dryad, Figshare, 4TU.ResearchData, and GitLab, among others. For most of these, an access token is optional, and only raises how many requests you may make per hour or gives you access to your own private material. For one of them, Dryad, a token is required even to download files from a fully public dataset.

The [Downloading an OSF Account](archiving-osf.qmd) and [Uploading Files to Zenodo](uploading-to-zenodo.qmd) chapters already explain, with screenshots, how to create and store a token for those two services. This chapter does the same for the remaining repositories: Dataverse, Dryad, Figshare, 4TU.ResearchData, and GitLab.

An access token is a long string of letters and numbers that proves to a website that a program is acting on your behalf. It works like a password that you give to one specific program, and that you can revoke without changing your real password.

Every token in this chapter is stored the same way: in a file called `.Renviron`, which R reads when it starts. Open it with:

```{r}
usethis::edit_r_environ()
```

Add the line for the service you need, using the exact name shown below, save the file, and restart R (**Ctrl+Shift+F10** in RStudio). The restart matters: `.Renviron` is only read when R starts, so a token added without restarting will not be found.

::: {.callout-note}
## Using a token for one session only

Every function below that reads a token (`dataverse_pat()`, `dryad_pat()`, `figshare_pat()`, `researchdata4tu_pat()`, `gitlab_pat()`) also sets it when you pass a value as an argument, for the rest of the current session only:

```{r}
figshare_pat("paste-your-token-here")
```

This is forgotten when you close R. Storing the token in `.Renviron` instead means you do not have to repeat this every time you start R.
:::

## Dataverse

Dataverse is not a single website but open-source software that many independent organisations each run on their own address — Harvard Dataverse, DataverseNL, and the DANS Data Stations are three examples. Because of this, **each Dataverse installation issues its own token**, and a token from one installation does nothing on another.

Go to the installation's own site and log in. Click your account name in the navigation bar, and choose **API Token** from the dropdown menu. This opens a page with a **Create Token** button (or **Recreate Token**, if you already have one). For Harvard Dataverse, for example, this account menu is reached from <https://dataverse.harvard.edu>.

Because the token is tied to one installation, storing it also requires naming the installation. `dataverse_pat()` takes a `host`, so store it under a name of your own choosing that includes the host, and pass that host when you check it:

```
DATAVERSE_HARVARD_PAT="paste-your-token-here"
```

```{r}
dataverse_pat(host = "dataverse.harvard.edu")
```

If you work with more than one Dataverse installation, add one line per installation, each with its own environment-variable name, and call `dataverse_pat()` with the matching `host` each time.

## Dryad

Dryad is a single service, at <https://datadryad.org>. Unlike every other repository in this chapter, a token is **not optional**: Dryad answers file-download requests with an error unless a token is present, even for a dataset anyone can see on the website. Without a token, you can still see what a dataset contains — Dryad's listing of a dataset's files works without one — but you cannot retrieve any of the files themselves.

### Creating a Dryad API account

Log in at <https://datadryad.org>, go to **My account**, and click **Create a Dryad API account**. This gives you an API account ID, a secret, and an initial access token.

::: {.callout-important}
## Every Dryad token expires after 10 hours

Unlike the other tokens in this chapter, a Dryad access token -- whether it is the initial one you were given, or one you obtain later -- is only valid for **10 hours**. After that it must be replaced with a new one. There is no way around this with a single stored token, which is why Dryad also gives you an account ID and secret: those do not expire, and can be exchanged for a fresh token automatically.
:::

### Using the initial token directly

For a single session, you can store the initial access token the same way as any other token in this chapter:

```
DRYAD_PAT="paste-your-token-here"
```

Check it:

```{r}
dryad_pat()
```

Once 10 hours have passed, this stops working, and you must sign in again, obtain a new token, and update `.Renviron`.

### Using the account ID and secret, for long-running work

Rather than storing a token that expires, you can give Metacheck your Dryad API account ID and secret. Metacheck then exchanges them for an access token itself, and repeats the exchange automatically whenever the token is about to expire, for as long as the account ID and secret remain valid:

```{r}
dryad_auth(
  client_id = "paste-your-api-account-id-here",
  client_secret = "paste-your-secret-here"
)
```

Unlike the token functions elsewhere in this chapter, `dryad_auth()` is normally called directly in your script rather than stored in `.Renviron`, because it takes two values rather than one. Use whichever of `dryad_pat()` or `dryad_auth()` suits the job: `dryad_pat()` for a quick session under 10 hours; `dryad_auth()` for a script that will keep running past that, such as building a large corpus of downloaded datasets.

## Figshare

Figshare is a single service at <https://figshare.com>, with several institutions — including the University of Leicester and DTU Data — running their own branded address on the same underlying service. A token is optional for reading public articles, where it only raises the rate limit; it is required to read a private article.

Create a token at <https://figshare.com/account/applications>. Log in, open that page, and create a new personal token there.

Store it as:

```
FIGSHARE_PAT="paste-your-token-here"
```

Check it:

```{r}
figshare_pat()
```

## 4TU.ResearchData

4TU.ResearchData, at <https://data.4tu.nl>, runs Djehuty, which is a documented copy of Figshare's own software. Because of this, its functions in Metacheck reuse Figshare's own code. This does **not** mean a Figshare.com token works here: 4TU.ResearchData is a separate organisation with its own accounts, so it issues its own tokens, stored separately from `figshare_pat()`.

Log in at <https://data.4tu.nl>, go to your **Applications**, and click **Create Personal Token**. Give it a short description, such as `metacheck`, and save; the token is shown once, so copy it before leaving the page.

Store it as:

```
RESEARCHDATA4TU_PAT="paste-your-token-here"
```

Check it:

```{r}
researchdata4tu_pat()
```

## GitLab

This covers gitlab.com itself, not a self-hosted GitLab instance run by another organisation. A token is optional for listing or downloading a public project, where it only raises GitLab's unauthenticated rate limit; it is required for a private project.

Log in at <https://gitlab.com>, open your avatar menu in the top right, choose **Edit profile**, and select **Access Tokens** in the left sidebar (or go directly to <https://gitlab.com/-/user_settings/personal_access_tokens>). Give the token a name and an expiry date, tick the **read_repository** scope, which covers reading a project's files over the repository files API, and click **Create personal access token**.

Store it as:

```
GITLAB_PAT="paste-your-token-here"
```

Check it:

```{r}
gitlab_pat()
```

## When something goes wrong

**A `*_pat()` function returns `""`.** R cannot find your token. Restart R, then check `.Renviron` has the line with quotation marks and a blank line at the end of the file, exactly as described in the [Downloading an OSF Account](archiving-osf.qmd#storing-the-osf-token) chapter — the same cause and the same fix apply to every token in this chapter.

**Dryad downloads fail with files listed but none retrieved.** This is expected without a token: Dryad allows listing a public dataset's files without one, but refuses to send the file contents. Set `DRYAD_PAT`, or use `dryad_auth()`, and try again.

**A Dryad token stops working after it used to work.** Every Dryad token expires after 10 hours, including the initial one from account creation. If you are storing it with `DRYAD_PAT`, sign in again and get a new one. Switch to `dryad_auth()` with your account ID and secret so this stops happening.

**A Dataverse token works on one installation but not another.** This is expected: each Dataverse installation issues and checks its own tokens. Create a separate token on the installation that is failing, and pass its `host` to `dataverse_pat()`.

**A 4TU.ResearchData request fails even though the Figshare token is set.** 4TU.ResearchData does not accept a Figshare.com token. Create a separate token on data.4tu.nl and store it as `RESEARCHDATA4TU_PAT`.
