---
title: "Checking Statistical Reporting"
vignette: >
  %\VignetteIndexEntry{Checking Statistical Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(metacheck)
```


## Journal Article Reporting Standards

The Journal Article Reporting Standards (JARS) developed by the American Psychological Association provide guidance on what information should be reported in scientific articles to support transparency and reproducibility [@appelbaum2018]. Almost every published article would improve if researchers diligently followed JARS, but the guidelines are not widely known, rarely enforced by journals, and time-consuming to check manually. Automation can help.

Metacheck provides five modules for checking statistical reporting quality. This vignette demonstrates each one using papers from the `psychsci` built-in dataset — 250 open-access articles published in *Psychological Science* between 2004 and 2014.

| Module | What it checks |
|---|---|
| `stat_p_exact` | Flags imprecise p-values (e.g., *p* < .05 or *p* = .000) |
| `stat_p_nonsig` | Flags non-significant p-values that may be misinterpreted |
| `stat_effect_size` | Checks whether t-tests and F-tests include an effect size |
| `stat_check` | Checks that p-values are mathematically consistent with the reported test statistic |
| `marginal` | Detects language like "marginally significant" or "approaching significance" |

All modules return a list with `$traffic_light` ("green", "yellow", "red", or "na"), `$summary_text`, and `$table` with detailed results.

---

## A paper where everything looks fine

We start with a paper that passes all checks. Paper `0956797613520608` has 6 detected p-values, all exact, all t-tests and F-tests paired with effect sizes, and no consistency errors flagged by StatCheck.

```{r}
paper_clean <- psychsci[["0956797613520608"]]
```

```{r}
module_run(paper_clean, "all_p_values")$summary_text
```

```{r}
module_run(paper_clean, "stat_p_exact")$traffic_light
```

```{r}
module_run(paper_clean, "stat_effect_size")$traffic_light
```

```{r}
module_run(paper_clean, "stat_check")$traffic_light
```


A green result means the module found no issues. This does not guarantee the paper is free of reporting problems — the modules use regular expressions and have known false negative rates — but it is a good sign.

---

## Exact p-values (`stat_p_exact`)

The APA Manual states:

> Report exact p values (e.g., *p* = .031) to two or three decimal places. However, report p values less than .001 as *p* < .001.

Exact p-values matter because they allow readers to include results in p-curve or z-curve analyses [@simonsohn2014; @bartos2020], and they enable tools like StatCheck to verify internal consistency. The `stat_p_exact` module flags any p-value reported as *p* < some threshold greater than .001 (e.g., *p* < .05) or as exactly zero (e.g., *p* = .000, which is mathematically impossible).

Paper `0956797619842261` has 60 detected p-values, 59 of which are exact. One is imprecise:

```{r}
paper_imprecise <- psychsci[["0956797619842261"]]
result_exact <- module_run(paper_imprecise, "stat_p_exact")

result_exact$traffic_light
result_exact$summary_text
```

```{r}
# The flagged p-value and its context
result_exact$table[result_exact$table$p_comp != "=", c("text", "p_comp", "p_value", "expanded")]
```


The module returns the sentence containing the imprecise p-value together with the surrounding sentence (the `expanded` column) to help assess the context.

---

## Missing effect sizes (`stat_effect_size`)

Reporting an effect size alongside every test result is a core JARS requirement. It helps readers assess practical significance and supports meta-analyses. The `stat_effect_size` module searches for t-tests and F-tests reported in APA format and checks whether an effect size metric (Cohen's *d*, *r*, partial η², *f*, etc.) appears in the same sentence.

The module has been validated on papers from Psychological Science: it detected 100% of t-tests with effect sizes, 99% of t-tests without, 99% of F-tests with, and 100% of F-tests without.

The same paper as above (`0956797619842261`) also has one test without an effect size:

```{r}
result_es <- module_run(paper_imprecise, "stat_effect_size")

result_es$traffic_light
result_es$summary_text
```

```{r}
# The test that is missing an effect size
result_es$table[result_es$table$es == FALSE, c("text", "test")]
```


A clean paper for comparison — all tests have effect sizes:

```{r}
module_run(paper_clean, "stat_effect_size")$summary_text
```


---

## Non-significant p-values (`stat_p_nonsig`)

Non-significant results are frequently misinterpreted as evidence that an effect is absent, when in fact they only indicate that the data were insufficient to reject the null hypothesis. The correct approach is to perform an equivalence test against a smallest effect size of interest [@lakens2018].

The `stat_p_nonsig` module returns all sentences containing non-significant p-values (p ≥ .05). It does not automatically judge the interpretation — it surfaces the sentences so researchers can review them. Paper `0956797618772822` has 28 such sentences:

```{r}
paper_nonsig <- psychsci[["0956797618772822"]]
result_nonsig <- module_run(paper_nonsig, "stat_p_nonsig")

result_nonsig$traffic_light
result_nonsig$summary_text
```

```{r}
# A sample of sentences with non-significant p-values
head(result_nonsig$table[, c("text", "p_value")], 3)
```


---

## Marginal significance (`marginal`)

Describing a result as "marginally significant", "approaching significance", or "trending towards significance" implies a gradient of evidence where none exists under null hypothesis significance testing. The `marginal` module detects these phrases.

The same paper `0956797619842261` contains two such sentences:

```{r}
result_marginal <- module_run(paper_imprecise, "marginal")

result_marginal$traffic_light
result_marginal$summary_text
```

```{r}
result_marginal$table[, "text"]
```


---

## Statistical consistency (StatCheck, `stat_check`)

StatCheck [@nuijten2016] recalculates the p-value from the reported test statistic and degrees of freedom and checks whether the reported p-value is consistent with the recalculation. It supports t-tests and F-tests in APA format.

**Important caveat:** StatCheck has a known false positive rate of approximately 43% — that is, roughly 43 out of every 100 flagged tests are not actually errors. Use the module to identify tests that warrant a closer look, not as definitive proof of an error. See the [StatCheck preprint](https://osf.io/preprints/psyarxiv/tcxaj_v1/) for full validation details.

The clean paper returns green:

```{r}
module_run(paper_clean, "stat_check")$summary_text
```


Paper `09567976231223130` triggers multiple flags:

```{r}
paper_statcheck <- psychsci[["09567976231223130"]]
result_sc <- module_run(paper_statcheck, "stat_check")

result_sc$traffic_light
result_sc$summary_text
```

```{r}
# Rows where an error was flagged — verify manually before concluding anything
errors <- result_sc$table[result_sc$table$error == TRUE,
                          c("raw", "reported_p", "computed_p", "decision_error")]
head(errors, 3)
```


The `decision_error` column is `TRUE` when the inconsistency changes the significance conclusion (i.e., the test would be non-significant if the computed p-value were used instead of the reported one, or vice versa). These are higher-priority cases to check manually.

---

## Running all modules at once

You can run multiple modules in sequence by passing the output of one module as the input to the next. The `summary_table` accumulates across modules:

```{r}
#| eval: false
result <- psychsci[["0956797619842261"]] |>
  module_run("stat_p_exact") |>
  module_run("stat_effect_size") |>
  module_run("stat_check") |>
  module_run("marginal") |>
  module_run("stat_p_nonsig")

result$summary_table
```


---

## References

Appelbaum, M., Cooper, H., Kline, R. B., Mayo-Wilson, E., Nezu, A. M., & Rao, S. M. (2018). Journal article reporting standards for quantitative research in psychology. *American Psychologist*, *73*(1), 3–25. <https://doi.org/10.1037/amp0000191>

Bartoš, F., & Schimmack, U. (2020). Z-curve 2.0: Estimating replication rates and discovery rates. *Meta-Psychology*. <https://doi.org/10.15626/MP.2021.2720>

Lakens, D. (2018). Equivalence tests: A practical primer for t-tests, correlations, and meta-analyses. *Social Psychological and Personality Science*, *8*(4), 355–362. <https://doi.org/10.1177/1948550617697177>

Nuijten, M. B., Hartgerink, C. H. J., van Assen, M. A. L. M., Epskamp, S., & Wicherts, J. M. (2016). The prevalence of statistical reporting errors in psychology (1985–2013). *Behavior Research Methods*, *48*(4), 1205–1226. <https://doi.org/10.3758/s13428-015-0664-2>

Simonsohn, U., Nelson, L. D., & Simmons, J. P. (2014). p-Curve and effect size. *Perspectives on Psychological Science*, *9*(6), 666–681. <https://doi.org/10.1177/1745691614553988>
